---
title: "MIDUS SFI Analysis"
author: "Charlie Collinge"
date: "2026-01-09"
output: html_document
---

# Set up

```{r}
knitr::opts_chunk$set(echo = TRUE)

#library packages#
library(readxl)
library(haven)
library(tidyverse)
library(openxlsx)
library(corrplot)
library(RColorBrewer)
library(pROC)
library(PRROC)
library(boot)
library(iml)
library(cocor)
library(nptest)
library(rms)


#specify an out directory for results & visualization#
outdir <- "/example/directory/"

#####################
#load initial data#
#####################
#import individual files#
maind <- read_excel("/sourcedirectory/M2 MR1 SFI data.xlsx", col_names = TRUE, guess_max = 12000)
mainv <- read_excel("/sourcedirectory/M2 MR1 SFI criterion.xlsx", col_names = TRUE, guess_max = 12000)
m2mke <- read_sav("/sourcedirectory/22840-0001-Data-REST.sav")
mR1mke <- read_sav("/sourcedirectory/36722-0001-Data-REST.sav")

#"maind" contains a set variables from MIDUS necessary to compute the SFI-10#
#(& by extension the SFI-8, and models based on age & sex)#
#"maindv" contains criterion variables used in this analysis#
#Data from the Milwaukee (mke) samples is offered by MIDUS as a#
#restricted data product that qualified researchers can apply for.#
#Data from the mke samples in M2 ("m2mke") and MR1 ("mR1mke")#
#are combined here with main sample data to create a final analytic dataset#

#Specific variables used can be ascertained from the code below.#
#However, a comprehensive list of variable names used in this analysis#
#will be provided upon request to: collinge@umn.edu#
  
head(maind)
head(mainv)
head(m2mke)
head(mR1mke)

#####################
#tidy#
#####################
dv <- merge(maind, mainv, by = c("MIDUSID", "SAMPLMAJ", "M2ID", "MRID"), all = TRUE)
head(dv)

dv$M2ID <- as.numeric(dv$M2ID)
dv$MRID <- as.numeric(dv$MRID)

#reconstructing MR1 iadl summary var (RA1SMADL)#
R1iadlqs <- c("RA1SA24A", "RA1SA24C", "RA1SA24E", "RA1SA24F", 
              "RA1SA24G", "RA1SA24I", "RA1SA24J")

for (nomiss in R1iadlqs) {
  x <- as.numeric(dv[[nomiss]])
  valid <- x %in% 1:4
  x[valid] <- 5 - x[valid]
  dv[[nomiss]] <- x
}

dv$RA1SMADL <- apply(dv[R1iadlqs], 1,
                     function(row) {
                       if (all(row == -1, na.rm = TRUE)) {return(-1)
                         }
                       valid <- row %in% 1:4
                       if (any(valid)) {
                         return(mean(row[valid]))
                       } else {return(8)
                       }
                       }
                     )
  
summary(dv$RA1SMADL[!(dv$RA1SMADL %in% c(-1, 8))])
table(!is.na(dv$RA1SMADL[!(dv$RA1SMADL %in% c(-1, 8))]))

dv$RA1SMADL[dv$RA1SMADL == -1] <- NA
dv$RA1SMADL[dv$RA1SMADL == 8] <- NA

colnames(mR1mke)[colnames(mR1mke) == "RAACIADL"] <- "RAACMADL"

#integrate MKE data#
m2mkevars <- setdiff(intersect(names(dv), names(m2mke)), c("M2ID", "SAMPLMAJ"))
mR1mkevars <- setdiff(intersect(names(dv), names(mR1mke)), c("MRID", "SAMPLMAJ"))

dv <- dv %>%
  left_join(m2mke, by = c("M2ID", "SAMPLMAJ"), suffix = c("", ".m2mke")) %>%
  mutate(across(all_of(m2mkevars), ~ get(paste0(cur_column(), ".m2mke")))) %>%
  select(-ends_with(".m2mke"))

dv <- dv %>%
  left_join(mR1mke, by = c("MRID", "SAMPLMAJ"), suffix = c("", ".mR1mke")) %>%
  mutate(across(all_of(mR1mkevars), ~ get(paste0(cur_column(), ".mR1mke")))) %>%
  select(-ends_with(".mR1mke"))

#combining core and MKE vars#
pairsdv <- list(
  interview_M_M2 = c("B1PIDATE_MO", "BACIDATE_MO"),
  interview_Y_M2 = c("B1PIDATE_YR", "BACIDATE_YR"),
  sex_M2 = c("B1PRSEX", "BACRSEX"),
  age_M2 = c("B1PRAGE_2019", "BACRAGE_2018"), 
  neighborhood_clean_M2 = c("B1SI6K", "BACIS6K"),
  financial_control_M2 = c("B1SG4", "BACGS4"), 
  family_contact_M2 = c("B1SJ1", "BACJS1"),
  friends_contact_M2 = c("B1SJ3", "BACJS3"),
  work_M2 = c("B1PBWORK", "BACB3WK"), 
  volunteerhn_M2 = c("B1SH7A", "BACHS7A"),
  volunteersoy_M2 = c("B1SH7B", "BACHS7B"),
  volunteerpoc_M2 = c("B1SH7C", "BACHS7C"),
  volunteerother_M2 = c("B1SH7D", "BACHS7D"),
  lonely_M2 = c("B1SA24G", "BACAS24G"),
  close_M2 = c("B1SA26G", "BACAS26G"),
  belong_M2 = c("B1SA26H", "BACAS26H"),
  courtesy_M2 = c("B1SP2A", "BACPS2A"),
  respect_M2 = c("B1SP2B", "BACPS2B"),
  interview_M_R1 = c("RA1PIDATE_MO", "RAACIDATE_MO"),
  interview_Y_R1 = c("RA1PIDATE_YR", "RAACIDATE_YR"),
  sex_R1 = c("RA1PRSEX", "RAACRSEX"),
  age_R1 = c("RA1PRAGE", "RAACRAGE"), 
  neighborhood_clean_R1 = c("RA1SJ6K", "RAACIS6K"),
  financial_control_R1 = c("RA1SH5", "RAACGS4"), 
  family_contact_R1 = c("RA1SK1", "RAASF1"),
  friends_contact_R1 = c("RA1SK5", "RAASF5"),
  work_R1 = c("RA1PB3WK", "RAACB3WK"), 
  volunteerhn_R1 = c("RA1SI7A", "RAACHS7A"),
  volunteersoy_R1 = c("RA1SI7B", "RAACHS7B"),
  volunteerpoc_R1 = c("RA1SI7C", "RAACHS7C"),
  volunteerother_R1 = c("RA1SI7D", "RAACHS7D"),
  lonely_R1 = c("RA1SA20G", "RAACAS24G"),
  close_R1 = c("RA1SA22G", "RAACAS26G"),
  belong_R1 = c("RA1SA22H", "RAACAS26H"),
  courtesy_R1 = c("RA1SQ2A", "RAACPS2A"),
  respect_R1 = c("RA1SQ2B", "RAACPS2B"), 
  health_M2 = c("B1PA1", "BACA1"),
  mehealth_M2 = c("B1PA2", "BACA2"),
  pwb_autonomy_M2 = c("B1SPWBA2", "BASPWBA2"),
  pwb_environment_M2 = c("B1SPWBE2", "BASPWBE2"), 
  pwb_growth_M2 = c("B1SPWBG2", "BASPWBG2"), 
  pwb_relations_M2 = c("B1SPWBR2", "BASPWBR2"),
  pwb_purpose_M2 = c("B1SPWBU2", "BASPWBU2"), 
  pwb_acceptance_M2 = c("B1SPWBS2", "BASPWBS2"), 
  swb_meaningful_M2 = c("B1SSWBMS", "BASSWBMS"), 
  swb_integration_M2 = c("B1SSWBSI", "BASSWBSI"), 
  swb_acceptance_M2 = c("B1SSWBAO", "BASSWBAO"), 
  swb_contribtuion_M2 = c("B1SSWBSC", "BASSWBSC"),
  swb_actualization_M2 = c("B1SSWBSA", "BASSWBSA"), 
  badl_M2 = c("B1SBADL2", "BACBADL2"),
  iadl_M2 = c("B1SMADL", "BACMADL"),
  mpq_wellbeing_M2 = c("B1SMPQWB", "BASMPQWB"), 
  mpq_socpotency_M2 = c("B1SMPQSP", "BASMPQSP"), 
  mpq_achieve_M2 = c("B1SMPQAC", "BASMPQAC"), 
  mpq_socclose_M2 = c("B1SMPQSC", "BASMPQSC"), 
  mpq_stressreact_M2 = c("B1SMPQSR", "BASMPQSR"), 
  mpq_aggression_M2 = c("B1SMPQAG", "BASMPQAG"), 
  mpq_alienation_M2 = c("B1SMPQAL", "BASMPQAL"), 
  mpq_control_M2 = c("B1SMPQCN", "BASMPQCN"), 
  mpq_tradition_M2 = c("B1SMPQTR", "BASMPQTR"),
  mpq_harmavoid_M2 = c("B1SMPQHA", "BASMPQHA"), 
  fmly_affectsolidarity_M2 = c("B1SFAMSO", "BACFAMSO"), 
  frnd_affectsolidarity_M2 = c("B1SFDSOL", "BACFDSOL"), 
  affneg_M2 = c("B1SNEGAF", "BACNEGAF"),
  affpos_M2 = c("B1SPOSAF", "BACPOSAF"), 
  health_R1 = c("RA1PA1", "RAACA1"),
  mehealth_R1 = c("RA1PA2", "RAACA2"),
  pwb_autonomy_R1 = c("RA1SPWBA2", "RAASPWBA2"),
  pwb_environment_R1 = c("RA1SPWBE2", "RAASPWBE2"), 
  pwb_growth_R1 = c("RA1SPWBG2", "RAASPWBG2"), 
  pwb_relations_R1 = c("RA1SPWBR2", "RAASPWBR2"),
  pwb_purpose_R1 = c("RA1SPWBU2", "RAASPWBU2"), 
  pwb_acceptance_R1 = c("RA1SPWBS2", "RAASPWBS2"), 
  swb_meaningful_R1 = c("RA1SSWBMS", "RAASSWBMS"), 
  swb_integration_R1 = c("RA1SSWBSI", "RAASSWBSI"), 
  swb_acceptance_R1 = c("RA1SSWBAO", "RAASSWBAO"), 
  swb_contribtuion_R1 = c("RA1SSWBSC", "RAASSWBSC"),
  swb_actualization_R1 = c("RA1SSWBSA", "RAASSWBSA"), 
  badl_R1 = c("RA1SBADL2", "RAACBADL2"),
  iadl_R1 = c("RA1SMADL", "RAACMADL"),
  mpq_wellbeing_R1 = c("RA1SMPQWB", "RAASMPQWB"), 
  mpq_socpotency_R1 = c("RA1SMPQSP", "RAASMPQSP"), 
  mpq_achieve_R1 = c("RA1SMPQAC", "RAASMPQAC"), 
  mpq_socclose_R1 = c("RA1SMPQSC", "RAASMPQSC"), 
  mpq_stressreact_R1 = c("RA1SMPQSR", "RAASMPQSR"), 
  mpq_aggression_R1 = c("RA1SMPQAG", "RAASMPQAG"), 
  mpq_alienation_R1 = c("RA1SMPQAL", "RAASMPQAL"), 
  mpq_control_R1 = c("RA1SMPQCN", "RAASMPQCN"), 
  mpq_tradition_R1 = c("RA1SMPQTR", "RAASMPQTR"),
  mpq_harmavoid_R1 = c("RA1SMPQHA", "RAASMPQHA"), 
  fmly_affectsolidarity_R1 = c("RA1SFAMSO", "RAASFAMSO"), 
  frnd_affectsolidarity_R1 = c("RA1SFDSOL", "RAASFDSOL"), 
  affneg_R1 = c("RA1SNEGAF", "RAACNEGAF"),
  affpos_R1 = c("RA1SPOSAF", "RAACPOSAF")
)

dv <- reduce(names(pairsdv), function(df, name) {
  vars <- pairsdv[[name]]
  if (all(vars %in% names(df))) {
    df <- mutate(df, !!name := coalesce(df[[vars[1]]], df[[vars[2]]]))
  }
  df
}, .init = dv)

head(dv)

dv <- dv[, c("MIDUSID", "M2ID", "MRID", "SAMPLMAJ", "DOD_M", "DOD_Y", names(pairsdv))]

#####################
#recoding & cleaning#
#M2 SFI-10 vars#
#####################
dv$neighborhood_clean_M2 <- ifelse(dv$neighborhood_clean_M2 == -1, NA, 
                                   ifelse(dv$neighborhood_clean_M2 == 8, NA, 
                                          dv$neighborhood_clean_M2))

dv$financial_control_M2 <- ifelse(dv$financial_control_M2 == 98, NA,
                                  ifelse(dv$financial_control_M2 == -1, NA,
                                         dv$financial_control_M2))

dv$family_contact_M2 <- ifelse(dv$family_contact_M2 == 98, NA,
                               ifelse(dv$family_contact_M2 == -1, NA, 
                                      dv$family_contact_M2))

dv$friends_contact_M2 <- ifelse(dv$friends_contact_M2 == 98, NA, 
                                ifelse(dv$friends_contact_M2 == -1, NA, 
                                       dv$friends_contact_M2)) 

#recode work to binary currently working/not working
dv$work_M2 <- ifelse(dv$work_M2 >= 97, NA, 
                     ifelse(dv$work_M2 %in% c(1, 2, 4, 9), 1, 0))

#create volunteer variable, excluding volunteering at school/with other youth#
volunteer <- names(dv)[grep("^volunteer", names(dv))]
list(volunteer)

dv[volunteer] <- lapply(dv[volunteer], function(x) ifelse(x == -1, NA, 
                                                          ifelse(x == 998, NA, x)))

dv$volunteer_M2 <- rowSums(dv[, c("volunteerhn_M2", "volunteerpoc_M2", "volunteerother_M2")], na.rm = FALSE)

#create feels isolated/loneliness#
lonelinesscols2 <- c("lonely_M2", "close_M2", "belong_M2")
dv[lonelinesscols2] <- lapply(dv[lonelinesscols2], function(x) ifelse(x == -1, NA, 
                                                                      ifelse(x == 8, NA, x)))

dv$lonely_M2 <- 6 - dv$lonely_M2

dv$loneliness_M2 <- rowSums(dv[, c("lonely_M2", "close_M2", "belong_M2")], na.rm = FALSE)

dv$loneliness_M2 <- dv$loneliness_M2 - 2
dv$loneliness_M2 <- 14 - dv$loneliness_M2

#create treated with less courtesy or respect than others#
court_resp2 <- c("courtesy_M2", "respect_M2")
dv[court_resp2] <- lapply(dv[court_resp2], function(x) ifelse(x == -1, NA, 
                                                              ifelse(x == 8, NA, x)))

dv$courtresp_M2 <- rowSums(dv[, c("courtesy_M2", "respect_M2")], na.rm = FALSE)

dv$courtresp_M2 <- dv$courtresp_M2 - 1

#####################
#recoding & cleaning#
#R1 SFI-10 vars#
#####################
dv$neighborhood_clean_R1 <- ifelse(dv$neighborhood_clean_R1 == -1, NA, 
                                   ifelse(dv$neighborhood_clean_R1 == 8, NA, 
                                          dv$neighborhood_clean_R1))

dv$financial_control_R1 <- ifelse(dv$financial_control_R1 == 98, NA,
                                  ifelse(dv$financial_control_R1 == -1, NA,
                                         dv$financial_control_R1))

dv$family_contact_R1 <- ifelse(dv$family_contact_R1 == 98, NA,
                               ifelse(dv$family_contact_R1 == -1, NA, 
                                      dv$family_contact_R1))

dv$friends_contact_R1 <- ifelse(dv$friends_contact_R1 == 98, NA, 
                                ifelse(dv$friends_contact_R1 == -1, NA, 
                                       dv$friends_contact_R1)) 

#recode work to binary currently working/not working
dv$work_R1 <- ifelse(dv$work_R1 >= 97, NA, 
                     ifelse(dv$work_R1 %in% c(1, 2, 4, 9), 1, 0))

#create volunteer variable, excluding volunteering at school/with other youth#
dv$volunteer_R1 <- rowSums(dv[, c("volunteerhn_R1", "volunteerpoc_R1", "volunteerother_R1")], na.rm = FALSE)

#create feels isolated/loneliness#
lonelinesscolsR <- c("lonely_R1", "close_R1", "belong_R1")
dv[lonelinesscolsR] <- lapply(dv[lonelinesscolsR], function(x) ifelse(x == -1, NA, 
                                                                      ifelse(x == 8, NA, x)))

dv$lonely_R1 <- 6 - dv$lonely_R1

dv$loneliness_R1 <- rowSums(dv[, c("lonely_R1", "close_R1", "belong_R1")], na.rm = FALSE)

dv$loneliness_R1 <- dv$loneliness_R1 - 2
dv$loneliness_R1 <- 14 - dv$loneliness_R1

#create treated with less courtesy or respect than others#
court_respR <- c("courtesy_R1", "respect_R1")
dv[court_respR] <- lapply(dv[court_respR], function(x) ifelse(x == -1, NA, 
                                                              ifelse(x == 8, NA, x)))

dv$courtresp_R1 <- rowSums(dv[, c("courtesy_R1", "respect_R1")], na.rm = FALSE)

dv$courtresp_R1 <- dv$courtresp_R1 - 1

#####################
#recoding & cleaning#
#criterion vars#
#####################
#physical & mental health#
dv$health_M2[dv$health_M2 == 7] <- NA
dv$mehealth_M2[dv$mehealth_M2 == 7] <- NA
dv$health_R1[dv$health_R1 == 7] <- NA
dv$mehealth_R1[dv$mehealth_R1 == 7] <- NA

dv$health_M2 <- 6 - dv$health_M2
dv$mehealth_M2 <- 6 - dv$mehealth_M2
dv$health_R1 <- 6 - dv$health_R1
dv$mehealth_R1 <- 6 - dv$mehealth_R1

#pwb scales#
pwb <- names(dv)[grep("^pwb", names(dv))]
list(pwb)

dv[pwb] <- lapply(dv[pwb], function(x) ifelse(x == -1, NA, 
                                                ifelse(x == 98, NA, x)))

#swb scales#
swb <- names(dv)[grep("^swb", names(dv))]
list(swb)

dv[swb] <- lapply(dv[swb], function(x) ifelse(x == -1, NA, 
                                              ifelse(x == 98, NA, x)))

#adls#
#basic#
dv$badl_M2[dv$badl_M2 == -1] <- NA
dv$badl_R1[dv$badl_R1 == -1] <- NA
dv$badl_M2[dv$badl_M2 == 8] <- NA
dv$badl_R1[dv$badl_R1 == 8] <- NA

#instrumental
dv$iadl_M2[dv$iadl_M2 == -1] <- NA
dv$iadl_R1[dv$iadl_M2 == -1] <- NA
dv$iadl_M2[dv$iadl_M2 == 8] <- NA
dv$iadl_R1[dv$iadl_M2 == 8] <- NA

#mpq scales#
mpq <- names(dv)[grep("^mpq", names(dv))]
list(mpq)

dv[mpq] <- lapply(dv[mpq], function(x) ifelse(x == -1, NA, 
                                              ifelse(x == 98, NA, x)))

#affect scales#
aff <- names(dv)[grep("^aff", names(dv))]
list(aff)

dv[aff] <- lapply(dv[aff], function(x) ifelse(x == -1, NA, 
                                              ifelse(x == 8, NA, x)))

#relationship quality scales#
#family#
fmly <- names(dv)[grep("^fmly", names(dv))]
list(fmly)

dv[fmly] <- lapply(dv[fmly], function(x) ifelse(x == -1, NA, 
                                              ifelse(x == 8, NA, x)))

#friends#
frnd <- names(dv)[grep("^frnd", names(dv))]
list(frnd)

dv[frnd] <- lapply(dv[frnd], function(x) ifelse(x == -1, NA, 
                                              ifelse(x == 8, NA, x)))

#####################
#further tidying#
#####################
#subset by sample#
M2 <- df[, grep("_M2$", names(df))]
M2 <- cbind(M2, df$MIDUSID, df$SAMPLMAJ)
colnames(M2)[colnames(M2) == "df$MIDUSID"] <- "MIDUSID"
colnames(M2)[colnames(M2) == "df$SAMPLMAJ"] <- "SAMPLMAJ"

R1 <- df[, grep("_R1$", names(df))]
R1 <- cbind(R1, df$MIDUSID, df$SAMPLMAJ)
colnames(R1)[colnames(R1) == "df$MIDUSID"] <- "MIDUSID"
colnames(R1)[colnames(R1) == "df$SAMPLMAJ"] <- "SAMPLMAJ"

#complete cases for M2#
pwbM2 <- names(M2)[grep("^pwb", names(M2))]
list(pwbM2)

swbM2 <- names(M2)[grep("^swb", names(M2))]
list(swbM2)

mpqM2 <- names(M2)[grep("^mpq", names(M2))]
list(mpqM2)

affM2 <- names(M2)[grep("^aff", names(M2))]
list(affM2)

frndM2 <- names(M2)[grep("^frnd", names(M2))]
list(frndM2)

fmlyM2 <- names(M2)[grep("^fmly", names(M2))]
list(fmlyM2)

dfM2 <- M2[complete.cases(M2[, c("health_M2", "mehealth_M2","badl_M2", 
                                 "iadl_M2", pwbM2, swbM2, mpqM2, affM2, 
                                 frndM2, fmlyM2)]), ]

dfM2 <- dfM2[dfM2$age_M2 >= 40 & dfM2$age_M2 <= 75, ]
view(dfM2)

#complete cases for R1#
pwbR1 <- names(R1)[grep("^pwb", names(R1))]
list(pwbR1)

swbR1 <- names(R1)[grep("^swb", names(R1))]
list(swbR1)

mpqR1 <- names(R1)[grep("^mpq", names(R1))]
list(mpqR1)

affR1 <- names(R1)[grep("^aff", names(R1))]
list(affR1)

frndR1 <- names(R1)[grep("^frnd", names(R1))]
list(frndR1)

fmlyR1 <- names(R1)[grep("^fmly", names(R1))]
list(fmlyR1)

dfR1 <- R1[complete.cases(R1[, c("health_R1", "mehealth_R1","badl_R1", 
                                 pwbR1, swbR1, mpqR1, affR1, 
                                 frndR1, fmlyR1)]), ]
dfR1 <- dfR1[dfR1$age_R1 >= 40 & dfR1$age_R1 <= 75, ]
view(dfR1)


#####################
#making 4 yr mortality binary#
#1 = yes, 0 = no variable#
#####################
#Add DOD#
dod <- df[, c(1, 5, 6)]
dfM2 <- merge(dfM2, dod, by = "MIDUSID", all.x = FALSE)
dfR1 <- merge(dfR1, dod, by = "MIDUSID", all.x = FALSE)

#basing 4yr mortality calculation on Shah et al.'s ACTUAL criteria (not 4yr mortality)#
#via Sachin: "We mentioned "4-year mortality" in the paper, but the way we defined the#
#death_flag (the outcome, a binary indicator) was to include all deaths that occurred by#
#the end of 2015 (Dec 31, 2015) for the 2010 cohort.#

#extrapolating this to M2 and R1#

#M2#
#counting all deaths before December 31st, 2009#

#R1 consisted of two samples: MRY, MRO#
#MRY = ages 25-54; conducted interviews between November 2011 and October 2012#
#MRO = ages 55-75; conducted interviews between July 2013 and May 2014#
#As these two samples were interviewed at different times, need to apply#
#different mortality date cutoffs based on Sachin's criteria for each sample#
#MRY counting all deaths before December 31st, 2016#
#MRO counting all deaths before December 31st, 2018#

#note: dates cannot be narrowed down from months#
#note: <= 2010 because December, 2009 would be 2009 + 1#
#in the formula below, so 2010. Same applies for others#

#dfM2#
dfM2$DOD <- dfM2$DOD_Y + dfM2$DOD_M/12
dfM2$mortality <- ifelse(is.na(dfM2$DOD), 0,
                         ifelse(dfM2$DOD <= 2010, 1, 0))

table(dfM2$mortality == 1)

#dfR1#
dfR1$DOD <- dfR1$DOD_Y + dfR1$DOD_M/12
dfR1$mortality <- ifelse(is.na(dfR1$DOD), 0,
                         ifelse(dfR1$age_R1 < 55 & dfR1$DOD <= 2017, 1,
                                ifelse(dfR1$age_R1 >= 55 & dfR1$DOD <= 2019, 1, 0)))

table(dfR1$mortality == 1)

#just checking when died#
dfM2$interview_M2 <- dfM2$interview_Y_M2 + dfM2$interview_M_M2/12
dfM2$mortalitytime <- dfM2$DOD - dfM2$interview_M2

summary(dfM2$mortalitytime[dfM2$mortality == 1])
hist(dfM2$mortalitytime[dfM2$mortality == 1], breaks = "sturges", freq = FALSE,
     xlim = c(0, 6), ylim = c(0, 0.35),
     main = "M2 Cohort: Mortality Time", xlab = "Time (Years)")
lines(density(dfM2$mortalitytime[dfM2$mortality == 1]), col = "#27933E")

dfR1$interview_R1 <- dfR1$interview_Y_R1 + dfR1$interview_M_R1/12
dfR1$mortalitytime <- dfR1$DOD - dfR1$interview_R1

summary(dfR1$mortalitytime[dfR1$mortality == 1])
hist(dfR1$mortalitytime[dfR1$mortality == 1], breaks = "sturges", freq = FALSE,
     xlim = c(0, 6), ylim = c(0, 0.35),
     main = "R1 Cohort: Mortality Time", xlab = "Time (Years)")
lines(density(dfR1$mortalitytime[dfR1$mortality == 1]), col = "#27933E")
```


# Descriptive stats - demographics, health status, function

The following 2 sections can be combined to recreate the descriptive statistics in Table S1.

### M2

```{r}
#merge in descriptive stats sheet#
desc <- read_excel("/sourcedirectory/M2 MR1 SFI Descriptive Stats.xlsx", col_names = TRUE, guess_max = 12000)

pairsdesc <- list(
  stroke_M2 = c("B1PA6A", "BACA6A"),
  cancer_M2 = c("B1PA26", "BACA26"), 
  brstcancer_M2 = c("B1PA28A", "BACA28A"),
  cervcancer_M2 = c("B1PA28B", "BACA28B"),
  colcancer_M2 = c("B1PA28C", "BACA28C"),
  lungcancer_M2 = c("B1PA28D", "BACA28D"),
  lympleuk_M2 = c("B1PA28E", "BACA28E"),
  ovarcancer_M2 = c("B1PA28F", "BACA28F"),
  prostcancer_M2 = c("B1PA28G", "BACA28G"),
  utercancer_M2 = c("B1PA28I", "BACA28I"),
  skincancer_M2 = c("B1PA28H", "BACA28H"),
  smokeev_M2 = c("B1PA38A", "BACA37"),
  smoken_M2 = c("B1PA39", "BACA39"),
  depress_M2 = c("B1PDEPAD", "BACDEPAD"),
  heartd_M2 = c("B1PA7", "BACA7"),
  diabetes_M2 = c("B1SA11X", "BACAS11X"),
  jointbone_M2 = c("B1SA11D", "BACAS11D"),
  race_M2 = c("B1PF7A", "BACF7A"),
  ethnicity_M2 = c("B1PF1", "BACF1"),
  educ_M2 = c("B1PB1", "BACB1"),
  hincome_M2 = c("B1STINC1", "BACTINC1"),
  mstat_M2 = c("B1PB19", "BACB19"),
  stroke_R1 = c("RA1PA6A", "RAACA6A"),
  cancer_R1 = c("RA1PA26", "RAACA26"), 
  brstcancer_R1 = c("RA1PA28A", "RAACA28A"), 
  cervcancer_R1 = c("RA1PA28B", "RAACA28B"),
  colcancer_R1 = c("RA1PA28C", "RAACA28C"),
  lungcancer_R1 = c("RA1PA28D", "RAACA28D"),
  lympleuk_R1 = c("RA1PA28E", "RAACA28E"),
  ovarcancer_R1 = c("RA1PA28F", "RAACA28F"),
  prostcancer_R1 = c("RA1PA28G", "RAACA28G"),
  utercancer_R1 = c("RA1PA28I", "RAACA28I"),
  skincancer_R1 = c("RA1PA28H", "RAACA28H"),
  smokeev_R1 = c("RA1PA38A", "RAACA37"),
  smoken_R1 = c("RA1PA39", "RAACA39"),
  depress_R1 = c("RA1PDEPAD", "RAACDEPAD"),
  heartd_R1 = c("RA1PA7", "RAACA7"),
  diabetes_R1 = c("RA1SA11X", "RAACAS11X"),
  jointbone_R1 = c("RA1SA11D", "RAACAS11D"),
  race_R1 = c("RA1PF7A", "RAACF7A"),
  ethnicity_R1 = c("RA1PF1", "RAACF1"),
  educ_R1 = c("RA1PB1", "RAACB1"),
  hincome_R1 = c("RA1STINC", "RAACTINC"),
  mstat_R1 = c("RA1PB19", "RAACB19")
)

desc <- reduce(names(pairsdesc), function(df, name) {
  vars <- pairsdesc[[name]]
  if (all(vars %in% names(df))) {
    df <- mutate(df, !!name := coalesce(df[[vars[1]]], df[[vars[2]]]))
  }
  df
}, .init = desc)

head(desc)

desc <- desc[, c("MIDUSID", "M2ID", "MRID", "SAMPLMAJ", names(pairsdesc))]

dfM2 <- merge(dfM2, desc, by = c("MIDUSID", "SAMPLMAJ"), all.x = TRUE)
view(dfM2)

#tidy#
cancerM2 <- c("brstcancer_M2", "cervcancer_M2", "colcancer_M2", "lungcancer_M2", 
              "lympleuk_M2", "ovarcancer_M2", "prostcancer_M2", "utercancer_M2")

dfM2[cancerM2] <- lapply(dfM2[cancerM2], function(x) ifelse(x == 7, NA,
                                                          ifelse(x == 8, NA, 
                                                              ifelse(x == 9, NA, x))))

dfM2$cancerflag_M2 <- ifelse((dfM2$cancer_M2 == 1) &
                                !(dfM2$skincancer_M2 == 1 & 
                                    rowSums(dfM2[cancerM2] == 1, na.rm = TRUE) == 0), 1, 2)

#####################
#demographics#
#####################
#age (median+IQR)#
summary(dfM2$age_M2)
hist(dfM2$age_M2, breaks = "sturges", freq = FALSE,
      main = "M2 Cohort: Age", xlab = "Years")
lines(density(dfM2$age_M2), col = "#27933E")

#sex (female+%)#
table(dfM2$sex_M2 == 2)

#####################
#health status#
#####################
#fair or poor self reported health (N+%)#
table(dfM2$health_M2 >= 4)

#diabetes (N+%)#
table(dfM2$diabetes_M2)

#cancer (excluding minor skin cancer) (N+%)#
table(dfM2$cancerflag_M2 == 1)

#heart disease (N+%)#
table(dfM2$heartd_M2 == 1)

#stroke (N+%)#
table(dfM2$stroke_M2 == 1)

#arthritis (N+%)#
table(dfM2$jointbone_M2)

#depression (MDE in past 12mo) (N+%)#
table(dfM2$depress_M2 == 1)

#hx of smoking (N+%)#
table(dfM2$smokeev_M2 == 1)

#current smoker? (N+%)#
table(dfM2$smoken_M2 == 1)

#####################
#mortality#
#####################
#mortality#
table(dfM2$mortality)


#####################
#summarize#
#####################
N <- 3542

freq_percent <- function(var, condition) {
  freq <- sum(var %in% condition, na.rm = TRUE)  
  perc <- (freq / N) * 100              
  return(paste0(freq, " (", round(perc, 2), "%)"))
}

freq_percentlog <- function(var, condition) {
  freq <- sum(condition, na.rm = TRUE)
  perc <- (freq / length(var)) * 100
  paste0(freq, " (", round(perc, 2), "%)")
}

M2stats <- data.frame(
  Variable = c(
    "Fair or Poor Self-Reported Health",
    "Diabetes",
    "Cancer (excluding minor skin cancer)",
    "Heart Disease",
    "Stroke",
    "Bone/Joint Disease",
    "Depression",
    "Current Smoker",
    "Former Smoker"
  ),
  `N (%)` = c(
    freq_percent(dfM2$health_M2, c(4, 5)), 
    freq_percent(dfM2$diabetes_M2, 1),  
    freq_percent(dfM2$cancerflag_M2, 1),  
    freq_percent(dfM2$heartd_M2, 1),  
    freq_percent(dfM2$stroke_M2, 1),  
    freq_percent(dfM2$jointbone_M2, 1),  
    freq_percent(dfM2$depress_M2, 1),  
    freq_percent(dfM2$smoken_M2, 1),  
    freq_percent(dfM2$smokeev_M2[dfM2$smoken_M2 == 2], 1)
  )
)

write.csv(M2stats, file.path(outdir, "M2 health and function.csv"), row.names = TRUE)
```

### R1

```{r}
dfR1 <- merge(dfR1, desc, by = c("MIDUSID", "SAMPLMAJ"), all.x = TRUE)
view(dfR1)

#tidy#
cancerR1 <- c("brstcancer_R1", "cervcancer_R1", "colcancer_R1", "lungcancer_R1", 
              "lympleuk_R1", "ovarcancer_R1", "prostcancer_R1", "utercancer_R1")

dfR1[cancerR1] <- lapply(dfR1[cancerR1], function(x) ifelse(x == 7, NA,
                                                          ifelse(x == 8, NA, 
                                                              ifelse(x == 9, NA, x))))

dfR1$cancerflag_R1 <- ifelse((dfR1$cancer_R1 == 1) &
                                !(dfR1$skincancer_R1 == 1 & 
                                    rowSums(dfR1[cancerR1] == 1, na.rm = TRUE) == 0), 1, 2)

#####################
#demographics#
#####################
#age (median+IQR)#
summary(dfR1$age_R1)
hist(dfR1$age_R1, breaks = "sturges", freq = FALSE, 
      main = "R1 Cohort: Age", xlab = "Years")
lines(density(dfR1$age_R1), col = "#27933E")

#sex (female+%)#
table(dfR1$sex_R1 == 2)

#####################
#health status#
#####################
#fair or poor self reported health (N+%)#
table(dfR1$health_R1 >= 4)

#diabetes (N+%)#
table(dfR1$diabetes_R1)

#cancer (excluding minor skin cancer) (N+%)#
table(dfR1$cancerflag_R1 == 1)

#heart disease (N+%)#
table(dfR1$heartd_R1 == 1)

#stroke (N+%)#
table(dfR1$stroke_R1 == 1)

#arthritis (N+%)#
table(dfR1$jointbone_R1)

#depression (MDE in past 12mo) (N+%)#
table(dfR1$depress_R1 == 1)

#hx of smoking (N+%)#
table(dfR1$smokeev_R1 == 1)

#current smoker? (N+%)#
table(dfR1$smoken_R1 == 1)

#####################
#mortality#
#####################
#mortality#
table(dfR1$mortality)


#####################
#summarize#
#####################
N <- 2077

R1stats <- data.frame(
  Variable = c(
    "Fair or Poor Self-Reported Health",
    "Diabetes",
    "Cancer (excluding minor skin cancer)",
    "Heart Disease",
    "Stroke",
    "Bone/Joint Disease",
    "Depression",
    "Current Smoker",
    "Former Smoker"
  ),
  `N (%)` = c(
    freq_percent(dfR1$health_R1, c(4, 5)), 
    freq_percent(dfR1$diabetes_R1, 1),  
    freq_percent(dfR1$cancerflag_R1, 1),  
    freq_percent(dfR1$heartd_R1, 1),  
    freq_percent(dfR1$stroke_R1, 1),  
    freq_percent(dfR1$jointbone_R1, 1),  
    freq_percent(dfR1$depress_R1, 1),  
    freq_percent(dfR1$smoken_R1, 1),  
    freq_percent(dfR1$smokeev_R1[dfR1$smoken_R1 == 2], 1)  
  )
)

write.csv(R1stats, file.path(outdir, "R1 health and function.csv"), row.names = TRUE)
```


# Descriptive stats - social characteristics

### M2

```{r}
#####################
#general resources#
#####################
#race (N+%)#
table(dfM2$race_M2)

#ethnicity (N+%)#
table(dfM2$ethnicity_M2)

#education (N+%)#
table(dfM2$educ_M2)

#income (N+%)#
summary(dfM2$hincome_M2)
dfM2$hincome_M2[dfM2$hincome_M2 == -1 ] <- NA
dfM2$hincome_M2[dfM2$hincome_M2 == 9999998 ] <- NA

summary(dfM2$hincome_M2)
hist(dfM2$hincome_M2, breaks = "sturges", freq = FALSE, 
     main = "M2 Cohort: Household Income", xlab = "Annual Income (USD)")
lines(density(dfM2$hincome_M2, na.rm = TRUE), col = "#27933E")

#marital status (N+%)#
table(dfM2$mstat_M2)

#####################
#sfi vars#
#####################
#all tables based on functional forms analysis in HRS#
#neighborhood unclean#
#code (< 3) to (0); (>= 3) to (1)#
table(dfM2$neighborhood_clean_M2 >= 3)

#low financial control#
#code (< 2) to (0); (>= 2) to (1)#
table(dfM2$financial_control_M2 < 2)

#low family contact#
#code (< 7) to (0); (>= 7) to (1)#
table(dfM2$family_contact_M2 >= 7)

#friend frequency#
summary(dfM2$friends_contact_M2)

#work#
table(dfM2$work_M2 == 1)

#no volunteer#
table(dfM2$volunteer_M2 == 0)

#loneliness#
table(dfM2$loneliness_M2)

#less courtesy/respect#
#code (< 5) to (0); (>= 5) to (1)#
table(dfM2$courtresp_M2 < 5)

#####################
#summarize#
#####################
N = 3542

M2descstats <- data.frame(
  Variable = c(
    "Black",
    "White",
    "Other", 
    "Hispanic origin", 
    "Not of hispanic origin", 
    "Less than high school", 
    "High school diploma or equiv", 
    "Some college", 
    "College and above",
    "< $56,599", 
    "$56,600 - 169,799", 
    "> $169,800", 
    "Age", 
    "Sex", 
    "Neighborhood is not clean", 
    "Low percieved financial control", 
    "Meets with children less than once per year or equivalent", 
    "Presently working", 
    "Does no volunteer/charity work", 
    "Feels isolated often", 
    "Feels isolated some of the time", 
    "Feels isolated hardly ever or never",
    "Often treated with less courtesy and respect than others"
  ),
  `N (%)` = c(
    freq_percent(dfM2$race_M2, 2),
    freq_percent(dfM2$race_M2, 1),
    freq_percent(dfM2$race_M2, 3:6),
    freq_percent(dfM2$ethnicity_M2, 2:7), 
    freq_percent(dfM2$ethnicity_M2, 1),
    freq_percent(dfM2$educ_M2, 1:3),
    freq_percent(dfM2$educ_M2, 4:5),
    freq_percent(dfM2$educ_M2, 6:7),
    freq_percent(dfM2$educ_M2, 8:12),
    freq_percentlog(dfM2$hincome_M2, dfM2$hincome_M2 < 56600),
    freq_percentlog(dfM2$hincome_M2, dfM2$hincome_M2 >= 56600 & dfM2$hincome_M2 <= 169799),
    freq_percentlog(dfM2$hincome_M2, dfM2$hincome_M2 > 169799),
    paste0(round(median(dfM2$age_M2), 2), 
           " [", round(summary(dfM2$age_M2)[2], 2),",", 
           round(summary(dfM2$age_M2)[5], 2), "]"), 
    freq_percent(dfM2$sex_M2, 2), 
    freq_percent(dfM2$neighborhood_clean_M2, 3:4), 
    freq_percent(dfM2$financial_control_M2, 0:1), 
    freq_percent(dfM2$family_contact_M2, 7:8), 
    #manually computed friends contact#
    freq_percent(dfM2$work_M2, 1), 
    freq_percent(dfM2$volunteer_M2, 0), 
    freq_percent(dfM2$loneliness_M2, 1:5), 
    freq_percent(dfM2$loneliness_M2, 6:10),
    freq_percent(dfM2$loneliness_M2, 11:13),
    freq_percent(dfM2$courtresp_M2, 1:4)
  )
)

write.csv(M2descstats, file.path(outdir, "M2 descriptive stats.csv"), row.names = TRUE)
```

### R1

```{r}
#####################
#general resources#
#####################
#race (N+%)#
table(dfR1$race_R1)

#ethnicity (N+%)#
table(dfR1$ethnicity_R1)

#education (N+%)#
table(dfR1$educ_R1)

#income (N+%)#
summary(dfR1$hincome_R1)
dfR1$hincome_R1[dfR1$hincome_R1 == -1 ] <- NA
dfR1$hincome_R1[dfR1$hincome_R1 == 999998 ] <- NA

summary(dfR1$hincome_R1)
hist(dfR1$hincome_R1, breaks = "sturges", freq = FALSE, 
     main = "R1 Cohort: Household Income", xlab = "Annual Income (USD)")
lines(density(dfR1$hincome_R1, na.rm = TRUE), col = "#27933E")


#marital status (N+%)#
table(dfR1$mstat_R1)

#####################
#sfi vars#
#####################
#all tables based on functional forms analysis in HRS#
#neighborhood unclean#
#code (< 3) to (0); (>= 3) to (1)#
table(dfR1$neighborhood_clean_R1 >= 3)

#low financial control#
#code (< 2) to (0); (>= 2) to (1)#
table(dfR1$financial_control_R1 < 2)

#low family contact#
#code (< 7) to (0); (>= 7) to (1)#
table(dfR1$family_contact_R1 >= 7)

#friend frequency#
summary(dfR1$friends_contact_R1)

#work#
table(dfR1$work_R1 == 1)

#no volunteer#
table(dfR1$volunteer_R1 == 0)

#loneliness#
table(dfR1$loneliness_R1)

#less courtesy/respect#
#code (< 5) to (0); (>= 5) to (1)#
table(dfR1$courtresp_R1 < 5)

#####################
#summarize#
#####################
N = 2077

R1descstats <- data.frame(
  Variable = c(
    "Black",
    "White",
    "Other", 
    "Hispanic origin", 
    "Not of hispanic origin", 
    "Less than high school", 
    "High school diploma or equiv", 
    "Some college", 
    "College and above",
    "< $56,599", 
    "$56,600 - 169,799", 
    "> $169,800", 
    "Age", 
    "Sex", 
    "Neighborhood is not clean", 
    "Low percieved financial control", 
    "Meets with children less than once per year or equivalent", 
    "Presently working", 
    "Does no volunteer/charity work", 
    "Feels isolated often", 
    "Feels isolated some of the time", 
    "Feels isolated hardly ever or never",
    "Often treated with less courtesy and respect than others"
  ),
  `N (%)` = c(
    freq_percent(dfR1$race_R1, 2),
    freq_percent(dfR1$race_R1, 1),
    freq_percent(dfR1$race_R1, 3:6),
    freq_percent(dfR1$ethnicity_R1, 2:7), 
    freq_percent(dfR1$ethnicity_R1, 1),
    freq_percent(dfR1$educ_R1, 1:3),
    freq_percent(dfR1$educ_R1, 4:5),
    freq_percent(dfR1$educ_R1, 6:7),
    freq_percent(dfR1$educ_R1, 8:12),
    freq_percentlog(dfR1$hincome_R1, dfR1$hincome_R1 < 56600),
    freq_percentlog(dfR1$hincome_R1, dfR1$hincome_R1 >= 56600 & dfR1$hincome_R1 <= 169799),
    freq_percentlog(dfR1$hincome_R1, dfR1$hincome_R1 > 169799),
    paste0(round(median(dfR1$age_R1), 2), 
           " [", round(summary(dfR1$age_R1)[2], 2),",", 
           round(summary(dfR1$age_R1)[5], 2), "]"), 
    freq_percent(dfR1$sex_R1, 2), 
    freq_percent(dfR1$neighborhood_clean_R1, 3:4), 
    freq_percent(dfR1$financial_control_R1, 0:1), 
    freq_percent(dfR1$family_contact_R1, 7:8), 
    #manually computed friends contact#
    freq_percent(dfR1$work_R1, 1), 
    freq_percent(dfR1$volunteer_R1, 0), 
    freq_percent(dfR1$loneliness_R1, 1:5), 
    freq_percent(dfR1$loneliness_R1, 6:10),
    freq_percent(dfR1$loneliness_R1, 11:13),
    freq_percent(dfR1$courtresp_R1, 1:4)
  )
)

write.csv(R1descstats, file.path(outdir, "R1 descriptive stats.csv"), row.names = TRUE)
```


# Find median (continuous) & mode values

```{r}
findmode <- function(x) {
  freq_table <- table(x)
  mode <- as.numeric(names(freq_table[freq_table == max(freq_table)]))
  return(mode)
}

#######################
#age (age_X)#
#######################
median(dfM2$age_M2, na.rm = TRUE)
median(dfR1$age_R1, na.rm = TRUE)

#median M2: 55#
#median R1: 58#

#######################
#volunteer (volunteer_X)#
#######################
median(dfM2$volunteer_M2, na.rm = TRUE)
median(dfR1$volunteer_R1, na.rm = TRUE)

#median M2: 0#
#median R1: 0#

#######################
#sex (sex_X)#
#######################
findmode(dfM2$sex_M2)
findmode(dfR1$sex_R1)

#mode M2: 2 (female)#
#mode R1: 2 (female)#

#######################
#neighborhood clean#
#(neighborhood_clean_X)#
#######################
findmode(dfM2$neighborhood_clean_M2)
findmode(dfR1$neighborhood_clean_R1)

#mode M2: 1 (highest agreement with "this area is kept very clean")#
#mode R1: 1 (highest agreement with "this area is kept very clean")#

#######################
#financial control#
#(financial_control_X)#
#######################
findmode(dfM2$financial_control_M2)
findmode(dfR1$financial_control_R1)

#mode M2: 8 (0 = "no control at all", 10 = "very much control")#
#mode R1: 8 (0 = "no control at all", 10 = "very much control")#

#######################
#family contact#
#(family_contact_X)#
#######################
findmode(dfM2$family_contact_M2)
findmode(dfR1$family_contact_R1)

#mode M2: 3 (several times a week)#
#mode R1: 3 (several times a week)#

#######################
#friends contact#
#(friends_contact_X)#
#######################
findmode(dfM2$friends_contact_M2)
findmode(dfR1$friends_contact_R1)

#mode M2: 3 (several times a week)#
#mode R1: 3 (several times a week)#

#######################
#work#
#(work_X)#
#######################
findmode(dfM2$work_M2)
findmode(dfR1$work_R1)

#mode: 1 (working currently)#
#mode: 1 (working currently)#

#######################
#loneliness#
#(loneliness_X)#
#######################
findmode(dfM2$loneliness_M2)
findmode(dfR1$loneliness_R1)

#mode M2: 11 (lower - bc reverse coded here to match HRS = more lonely)#
#mode R1: 11 (lower - bc reverse coded here to match HRS = more lonely)#

#######################
#treated w/ less courtesy/respect#
#(courtresp_X)#
#######################
findmode(dfM2$courtresp_M2)
findmode(dfR1$courtresp_R1)

#mode M2: 7 (lower = experienced, higher = not experienced)#
#mode R1: 7 (lower = experienced, higher = not experienced)#
```


# Impute & recode

```{r}
##############################
#impute & recode#
##############################
#age#
table(!is.na(dfM2$age_M2))
table(!is.na(dfR1$age_R1))

#sex#
table(!is.na(dfM2$sex_M2))
table(!is.na(dfR1$sex_R1))

dfM2$sex_M2[dfM2$sex_M2 == 1] <- 0
dfM2$sex_M2[dfM2$sex_M2 == 2] <- 1

dfR1$sex_R1[dfR1$sex_R1 == 1] <- 0
dfR1$sex_R1[dfR1$sex_R1 == 2] <- 1

#neighborhood cleanliness#
table(!is.na(dfM2$neighborhood_clean_M2))
table(!is.na(dfR1$neighborhood_clean_R1))

dfM2$neighborhood_clean_M2[is.na(dfM2$neighborhood_clean_M2)] <- 1
dfR1$neighborhood_clean_R1[is.na(dfR1$neighborhood_clean_R1)] <- 1

table(!is.na(dfM2$neighborhood_clean_M2))
table(!is.na(dfR1$neighborhood_clean_R1))

#financial control#
table(!is.na(dfM2$financial_control_M2))
table(!is.na(dfR1$financial_control_R1))

dfM2$financial_control_M2[is.na(dfM2$financial_control_M2)] <- 8
dfR1$financial_control_R1[is.na(dfR1$financial_control_R1)] <- 8

table(!is.na(dfM2$financial_control_M2))
table(!is.na(dfR1$financial_control_R1))

#family contact frequency#
table(!is.na(dfM2$family_contact_M2))
table(!is.na(dfR1$family_contact_R1))

dfM2$family_contact_M2[is.na(dfM2$family_contact_M2)] <- 3
dfR1$family_contact_R1[is.na(dfR1$family_contact_R1)] <- 3

table(!is.na(dfM2$family_contact_M2))
table(!is.na(dfR1$family_contact_R1))

#friend contact frequency#
table(!is.na(dfM2$friends_contact_M2))
table(!is.na(dfR1$friends_contact_R1))

dfM2$friends_contact_M2[is.na(dfM2$friends_contact_M2)] <- 3
dfR1$friends_contact_R1[is.na(dfR1$friends_contact_R1)] <- 3

table(!is.na(dfM2$friends_contact_M2))
table(!is.na(dfR1$friends_contact_R1))

#work status#
table(!is.na(dfM2$work_M2))
table(!is.na(dfR1$work_R1))

dfM2$work_M2[is.na(dfM2$work_M2)] <- 1
dfR1$work_R1[is.na(dfR1$work_R1)] <- 1

table(!is.na(dfM2$work_M2))
table(!is.na(dfR1$work_R1))

#volunteer#
table(!is.na(dfM2$volunteer_M2))
table(!is.na(dfR1$volunteer_R1))

dfM2$volunteer_M2[is.na(dfM2$volunteer_M2)] <- 0
dfR1$volunteer_R1[is.na(dfR1$volunteer_R1)] <- 0

table(!is.na(dfM2$volunteer_M2))
table(!is.na(dfR1$volunteer_R1))

#feels isolated/loneliness#
table(!is.na(dfM2$loneliness_M2))
table(!is.na(dfR1$loneliness_R1))

dfM2$loneliness_M2[is.na(dfM2$loneliness_M2)] <- 11
dfR1$loneliness_R1[is.na(dfR1$loneliness_R1)] <- 11

table(!is.na(dfM2$loneliness_M2))
table(!is.na(dfR1$loneliness_R1))

#treated with less courtesy or respect than others#
table(!is.na(dfM2$courtresp_M2))
table(!is.na(dfR1$courtresp_R1))

dfM2$courtresp_M2[is.na(dfM2$courtresp_M2)] <- 7
dfR1$courtresp_R1[is.na(dfR1$courtresp_R1)] <- 7

table(!is.na(dfM2$courtresp_M2))
table(!is.na(dfR1$courtresp_R1))
```


# Recode vars to manual PNAS coding

```{r}
#######################
#create dummy df for og coding#
#######################
dfM2ff <- dfM2
dfM2og <- dfM2
dfR1oc <- dfR1

#neighborhood cleanliness#
dfM2ff$neighborhood_clean_M2 <- ifelse(dfM2ff$neighborhood_clean_M2 == 1, 0, 
                                ifelse(dfM2ff$neighborhood_clean_M2 == 2, 0, 1))


table(dfM2ff$neighborhood_clean_M2)

#financial control#
#using (< 2) to (0) and >= 2 to (1) here based on HRS findings#
dfM2ff$financial_control_M2 <- ifelse(dfM2ff$financial_control_M2 >= 2, 1, 0)

table(dfM2ff$financial_control_M2)

#family contact frequency#
dfM2ff$family_contact_M2 <- ifelse(dfM2ff$family_contact_M2 >= 7, 1, 0)

table(dfM2ff$family_contact_M2)

#friend contact frequency#
dfM2ff$friends_contact_M2 <-  ifelse(dfM2ff$friends_contact_M2 <= 2, 1, 
                                              dfM2ff$friends_contact_M2 - 1)

table(dfM2ff$friends_contact_M2)

#volunteer#
dfM2ff$volunteer_M2 <- ifelse(dfM2ff$volunteer_M2 == 0, 1, 0)

table(dfM2ff$volunteer_M2)

#feels isolated/loneliness#
#recode#
lonely0 <- c(1, 2, 3, 4, 5)
lonely1 <- c(6, 7, 8, 9, 10)
lonely2 <- c(11, 12, 13)

dfM2ff$loneliness_M2 <- ifelse(dfM2ff$loneliness_M2 %in% lonely0, 0, 
                                ifelse(dfM2ff$loneliness_M2 %in% lonely1, 1, 2))

table(dfM2ff$loneliness_M2)

#treated with less courtesy or respect than others#
#recode#
courtresp0 <- c(1, 2, 3, 4)
courtresp1 <- c(5, 6, 7)

dfM2ff$courtresp_M2 <- ifelse(dfM2ff$courtresp_M2 %in% courtresp0, 0, 1)

table(dfM2ff$courtresp_M2)

#inspect#
view(dfM2ff)
```


# Functional forms

```{r}
#######################
#var types#
#######################
continuousvars <- c("age_M2", "volunteer_M2")
ordinalvars <- c("neighborhood_clean_M2", "financial_control_M2", "family_contact_M2", "friends_contact_M2",
                 "loneliness_M2", "courtresp_M2")

#######################
#continuous#
#######################

########
#age_M2#
alin <- glm(mortality ~ age_M2, data = dfM2og, family = binomial, na.action = na.omit)

dfM2og$logage_M2 <- log(dfM2og$age_M2 + 1)
alog <- glm(mortality ~ logage_M2, data = dfM2og, family = binomial, na.action = na.omit)

dfM2og$expage_M2 <- exp(dfM2og$age_M2)
aexp <- glm(mortality ~ expage_M2, data = dfM2og, family = binomial, na.action = na.omit)

abic <- data.frame(
  Model = c("Linear", "Log", "Exp"),
  BIC = c(BIC(alin), BIC(alog), BIC(aexp))
)

print(abic)


########
#volunteer_M2#
vlin <- glm(mortality ~ volunteer_M2, data = dfM2og, family = binomial, na.action = na.omit)

dfM2og$logvolunteer_M2 <- log(dfM2og$volunteer_M2 + 1)
vlog <- glm(mortality ~ logvolunteer_M2, data = dfM2og, family = binomial, na.action = na.omit)

dfM2og$expvolunteer_M2 <- exp(dfM2og$volunteer_M2)
vexp <- glm(mortality ~ expvolunteer_M2, data = dfM2og, family = binomial, na.action = na.omit)

vbic <- data.frame(
  Model = c("Linear", "Log", "Exp"),
  BIC = c(BIC(vlin), BIC(vlog), BIC(vexp))
)

print(vbic)

#######################
#ordinal#
#######################

########
#neighborhood_clean_M2#
nclin <- glm(mortality ~ neighborhood_clean_M2, data = dfM2og, family = binomial, na.action = na.omit)

dfM2og[, "neighborhood_clean_M2"] <- as.factor(dfM2og[,"neighborhood_clean_M2"])
nccat <- glm(mortality ~ neighborhood_clean_M2, data = dfM2og, family = binomial, na.action = na.omit)

ncman <- glm(mortality ~ neighborhood_clean_M2, data = dfM2ff, family = binomial, na.action = na.omit)

ncbic <- data.frame(
  Model = c("Linear", "Categorical", "Binary"),
  BIC = c(BIC(nclin), BIC(nccat), BIC(ncman))
)

print(ncbic)

########
#financial_control_M2# 
fclin <- glm(mortality ~ financial_control_M2, data = dfM2og, family = binomial, na.action = na.omit)

dfM2og[, "financial_control_M2"] <- as.factor(dfM2og[,"financial_control_M2"])
fccat <- glm(mortality ~ financial_control_M2, data = dfM2og, family = binomial, na.action = na.omit)

fcman <- glm(mortality ~ financial_control_M2, data = dfM2ff, family = binomial, na.action = na.omit)

fcbic <- data.frame(
  Model = c("Linear", "Categorical", "Binary"),
  BIC = c(BIC(fclin), BIC(fccat), BIC(fcman))
)

print(fcbic)


########
#family_contact_M2# 
facolin <- glm(mortality ~ family_contact_M2, data = dfM2og, family = binomial, na.action = na.omit)

dfM2og[, "family_contact_M2"] <- as.factor(dfM2og[,"family_contact_M2"])
facocat <- glm(mortality ~ family_contact_M2, data = dfM2og, family = binomial, na.action = na.omit)

facoman <- glm(mortality ~ family_contact_M2, data = dfM2ff, family = binomial, na.action = na.omit)

facobic <- data.frame(
  Model = c("Linear", "Categorical", "Binary"),
  BIC = c(BIC(facolin), BIC(facocat), BIC(facoman))
)

print(facobic)


########
#friends_contact_M2#
frcolin <- glm(mortality ~ friends_contact_M2, data = dfM2og, family = binomial, na.action = na.omit)

dfM2og[, "friends_contact_M2"] <- as.factor(dfM2og[,"friends_contact_M2"])
frcocat <- glm(mortality ~ friends_contact_M2, data = dfM2og, family = binomial, na.action = na.omit)

frcoman <- glm(mortality ~ friends_contact_M2, data = dfM2ff, family = binomial, na.action = na.omit)

frcobic <- data.frame(
  Model = c("Linear", "Categorical", "Binary"),
  BIC = c(BIC(frcolin), BIC(frcocat), BIC(frcoman))
)

print(frcobic)


########
#loneliness_M2#
lolin <- glm(mortality ~ loneliness_M2, data = dfM2og, family = binomial, na.action = na.omit)

dfM2og[, "loneliness_M2"] <- as.factor(dfM2og[,"loneliness_M2"])
locat <- glm(mortality ~ loneliness_M2, data = dfM2og, family = binomial, na.action = na.omit)

loman <- glm(mortality ~ loneliness_M2, data = dfM2ff, family = binomial, na.action = na.omit)

lobic <- data.frame(
  Model = c("Linear", "Categorical", "Binary"),
  BIC = c(BIC(lolin), BIC(locat), BIC(loman))
)

print(lobic)


########
#courtresp_M2#
crlin <- glm(mortality ~ courtresp_M2, data = dfM2og, family = binomial, na.action = na.omit)

dfM2og[, "courtresp_M2"] <- as.factor(dfM2og[,"courtresp_M2"])
crcat <- glm(mortality ~ courtresp_M2, data = dfM2og, family = binomial, na.action = na.omit)

crman <- glm(mortality ~ courtresp_M2, data = dfM2ff, family = binomial, na.action = na.omit)

crbic <- data.frame(
  Model = c("Linear", "Categorical", "Binary"),
  BIC = c(BIC(crlin), BIC(crcat), BIC(crman))
)

print(crbic)

summary <- cbind(abic$BIC, vbic$BIC, ncbic$BIC, fcbic$BIC, facobic$BIC, frcobic$BIC,
                 lobic$BIC, crbic$BIC)
colnames(summary) <- c("age", "volunteer", "neighborhood cleanliness", "financial control", 
                       "family contact", "friends contact", "loneliness", "courtesy/respect")
rownames(summary) <- c("1", "2", "3")
summary <- as.data.frame(summary)
summary

write.csv(summary, file.path(outdir, "MIDUS functional forms.csv"), row.names = FALSE)
```


# Recode vars based on functional forms

```{r}
#######################
#age (age_M2)#
#######################
#check#
summary(dfM2$age_M2)

#no change#

#######################
#sex (sex_M2)#
#######################
#check#
table(dfM2$sex_M2)
dfM2$sex_M2 <- as.numeric(dfM2$sex_M2)

#no change#

#######################
#neighborhood clean?#
#(neighborhood_clean_M2)#
#######################
#check#
table(dfM2$neighborhood_clean_M2)

#code (1, 2) = (0), (3, 4) = (1)#
dfM2$neighborhood_clean_M2 <- ifelse(dfM2$neighborhood_clean_M2 == 1, 0, 
                                ifelse(dfM2$neighborhood_clean_M2 == 2, 0, 1))


#check#
table(dfM2$neighborhood_clean_M2)

#######################
#financial control#
#(financial_control_M2)#
#######################
#check#
table(dfM2$financial_control_M2)

#no change#

#######################
#family contact#
#(family_contact_M2)#
#######################
#check#
table(dfM2$family_contact_M2)

#code (<7) = (0), (7, 8) = 1#
dfM2$family_contact_M2 <- ifelse(dfM2$family_contact_M2 >= 7, 1, 0)

#check#
table(dfM2$family_contact_M2)

#######################
#friends contact#
#(friends_contact_M2)#
#######################
#check#
table(dfM2$friends_contact_M2)

#compress bottom (1, 2) = 1, subtract 1 from else#
dfM2$friends_contact_M2 <-  ifelse(dfM2$friends_contact_M2 <= 2, 1, 
                                              dfM2$friends_contact_M2 - 1)

#check#
table(dfM2$friends_contact_M2)

#######################
#work for pay?#
#(work_M2)#
#######################
#check#
table(dfM2$work_M2)

#no change#

#######################
#volunteer#
#(volunteer_M2)#
#######################
#check#
summary(dfM2$volunteer_M2)

#change to log scale#
dfM2$volunteer_M2 <- log(dfM2$volunteer_M2 + 1)

#check#
summary(dfM2$volunteer_M2)

#######################
#loneliness#
#(loneliness_M2)#
#######################
#check#
table(dfM2$loneliness_M2)

#no change#

#######################
#treated w/ less courtesy/respect#
#(courtresp_M2)#
#######################
#check#
table(dfM2$courtresp_M2)

#code (< 5) to (0); (>= 5) to (1)#
courtresp0 <- c(1, 2, 3, 4)
courtresp1 <- c(5, 6, 7)

dfM2$courtresp_M2 <- ifelse(dfM2$courtresp_M2 %in% courtresp0, 0, 1)

#check#
table(dfM2$courtresp_M2)
```


# Developing models in M2

### SFI-10

```{r}
#######################
#rename cols to be congruent#
#w HRS names#
#######################
colnames(dfM2)[c(5:11, 50:52)] <- c("sex", "age", "neighborhood", "financial_control", 
                             "freq_children", "activities_youth", "work", "volunteering", 
                             "feel_isolated", "courtresp")

#######################
#specifying models#
#######################
#SFI10#
sfi <- glm(mortality ~ age + sex + neighborhood + financial_control + freq_children 
           + activities_youth + work + volunteering + feel_isolated + courtresp, 
           family = binomial, data = dfM2, na.action = na.omit)

summary(sfi)

dfM2$SFIlp_M2 <- predict(sfi, type = "link")
dfM2$SFIpr_M2 <- predict(sfi, type = "response")

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "M2SFI10distributions.pdf"), width = 6, height = 5)
hist(dfM2$SFIlp_M2, breaks = "sturges", freq = FALSE, 
     xlim = c(-8, 2), ylim = c(0, 0.35), 
     main = "M2 Cohort: SFI-10", xlab = "Mortality Risk (Log Probability)")
lines(density(dfM2$SFIlp_M2), col = "#27933E")
hist(dfM2$SFIpr_M2, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 25), 
     main = "M2 Cohort: SFI-10", xlab = "Mortality Risk (Probability)")
lines(density(dfM2$SFIpr_M2), col = "#27933E")
dev.off()
```

### Age

```{r}
#######################
#specifying models#
#######################
glmage <- glm(mortality ~ age, family = binomial, data = dfM2, na.action = na.omit)
summary(glmage)

dfM2$agelp <- predict(glmage, type = "link")
dfM2$agepr <- predict(glmage, type = "response")

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "M2glmagedistributions.pdf"), width = 6, height = 5)
hist(dfM2$agelp, breaks = "sturges", freq = FALSE, 
     xlim = c(-6, -1), ylim = c(0, 0.5), 
     main = "M2 Cohort: Age", xlab = "Mortality Risk (Log Probability)")
lines(density(dfM2$agelp), col = "#27933E")
hist(dfM2$agepr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 25), 
     main = "M2 Cohort: Age", xlab = "Mortality Risk (Probability)")
lines(density(dfM2$agepr), col = "#27933E")
dev.off()
```

### Age & Sex

```{r}
#######################
#specifying models#
#######################
sexage <- glm(mortality ~ age + sex, family = binomial, data = dfM2, na.action = na.omit)
summary(sexage)

dfM2$sexagelp <- predict(sexage, type = "link")
dfM2$sexagepr <- predict(sexage, type = "response")

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "M2glmagesexdistributions.pdf"), width = 6, height = 5)
hist(dfM2$sexagelp, breaks = "sturges", freq = FALSE, 
     xlim = c(-6, -1), ylim = c(0, 0.4), 
     main = "M2 Cohort: Age & Sex", xlab = "Mortality Risk (Log Probability)")
lines(density(dfM2$sexagelp), col = "#27933E")
hist(dfM2$sexagepr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 30), 
     main = "M2 Cohort: Age & Sex", xlab = "Mortality Risk (Probability)")
lines(density(dfM2$sexagepr), col = "#27933E")
dev.off()
```

### SFI-8

```{r}
#######################
#specifying models#
#######################
sfisoc <- glm(mortality ~ neighborhood + financial_control + freq_children 
           + activities_youth + work + volunteering + feel_isolated + courtresp, 
           family = binomial, data = dfM2, na.action = na.omit)

summary(sfisoc)

dfM2$socSFIlp <- predict(sfisoc, type = "link")
dfM2$socSFIpr <- predict(sfisoc, type = "response")

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "M2SFI8distributions.pdf"), width = 6, height = 5)
hist(dfM2$socSFIlp, breaks = "sturges", freq = FALSE, 
     xlim = c(-6, -1), ylim = c(0, 1), 
     main = "M2 Cohort: SFI-8", xlab = "Mortality Risk (Log Probability)")
lines(density(dfM2$socSFIlp), col = "#27933E")
hist(dfM2$socSFIpr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 40), 
     main = "M2 Cohort: SFI-8", xlab = "Mortality Risk (Probability)")
lines(density(dfM2$socSFIpr), col = "#27933E")
dev.off()
```

### AIC/BIC

```{r}
#all models specified in R#
midusglmsBIC <- BIC(glmage, sexage, sfisoc, sfi)
midusglmsAIC <- AIC(glmage, sexage, sfisoc, sfi)

midusglmmodelfit <- cbind(midusglmsBIC, midusglmsAIC$AIC)

print(midusglmmodelfit)

write.csv(midusglmmodelfit, file.path(outdir, "M2ModelFit.csv"), row.names = TRUE)
```

### BCa bootstrap fucntions

```{r}
stratboot <- function(data, statistic, R = 9999, strata, conf.level = 0.95, ...) {
  levels <- unique(strata)
  n <- nrow(data)

  resamples <- replicate(R, {
    unlist(lapply(levels, function(lvl) {
      idx <- which(strata == lvl)
      sample(idx, length(idx), replace = TRUE)
    }))
  }, simplify = FALSE)

  statwrap <- function(data, indices) statistic(data[indices, ], ...)

  bootobj <- boot(data = data, statistic = statwrap, R = R, sim = "ordinary")
  bootobj$t <- sapply(resamples, function(i) statwrap(data, i))
  bootobj$t <- matrix(bootobj$t, ncol = 1)

  ci <- boot.ci(bootobj, type = "bca", conf = conf.level)

  list(
    estimate = statwrap(data, 1:n),
    bca_ci = ci$bca[4:5],
    bootobj = bootobj
  )
}

auroc <- function(data) {
  pROC::auc(pROC::roc(data$response, data$predicted, quiet = TRUE)) |> as.numeric()
}

auprc <- function(data) {
  PRROC::pr.curve(
    scores.class0 = data$predicted[data$response == 1],
    scores.class1 = data$predicted[data$response == 0],
    curve = FALSE)$auc.integral
}

aurocdiff <- function(data) {
  auc1 <- pROC::auc(pROC::roc(data$response, data$pred1, quiet = TRUE)) |> as.numeric()
  auc2 <- pROC::auc(pROC::roc(data$response, data$pred2, quiet = TRUE)) |> as.numeric()
  auc1 - auc2
}

auprcdiff <- function(data) {
  pr1 <- PRROC::pr.curve(scores.class0 = data$pred1[data$response == 1],
                         scores.class1 = data$pred1[data$response == 0],
                         curve = FALSE)$auc.integral
  pr2 <- PRROC::pr.curve(scores.class0 = data$pred2[data$response == 1],
                         scores.class1 = data$pred2[data$response == 0],
                         curve = FALSE)$auc.integral
  pr1 - pr2
}
```

### M2 AUCs

```{r}
#######################
#SFI-10#
#######################
set.seed(555)
roccurvesfiM2 <- roc(dfM2$mortality, dfM2$SFIpr_M2, auc = TRUE, 
                     ci = TRUE, of = "auc", ci.method = "bootstrap", 
                     conf.level = 0.95, boot.n = 9999, boot.stratified = TRUE)
aucsfiM2 <- roccurvesfiM2$auc
ciaucsfiM2 <-  roccurvesfiM2$ci
print(aucsfiM2)
print(ciaucsfiM2)

dfbcarocsfiM2 <- data.frame(response = dfM2$mortality, 
                            predicted = dfM2$SFIpr_M2)
bcaciaucsfiM2 <- stratboot(dfbcarocsfiM2, auroc, R = 9999, strata = dfbcarocsfiM2$response)
print(bcaciaucsfiM2)

dfrocsfiM2 <- data.frame(specificity = 1 - roccurvesfiM2$specificities, 
                         sensitivity = roccurvesfiM2$sensitivities)

pdf(file.path(outdir, "M2AUCROC.pdf"), width = 5, height = 5)
ggplot(dfrocsfiM2, aes(x = specificity, y = sensitivity)) + 
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
    labs(title = paste0("M2 SFI-10: Mortality (AUC = ", round(aucsfiM2, 3), ")"), 
         x = "1 - Specificity", y = "Sensitivity") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))
dev.off()

prcurvesfiM2 <- pr.curve(scores.class0 = dfM2$SFIpr_M2, weights.class0 = dfM2$mortality, curve = TRUE)
auprcsfiM2 <- prcurvesfiM2$auc.integral
print(auprcsfiM2)

dfbcaprcsfiM2 <- data.frame(response = dfM2$mortality, 
                            predicted = dfM2$SFIpr_M2)
bcaciauprcsfiM2 <- stratboot(dfbcarocsfiM2, auprc, R = 9999, strata = dfbcaprcsfiM2$response)
print(bcaciauprcsfiM2)

dfprcsfiM2 <- data.frame(recall = prcurvesfiM2$curve[, 1], 
                         precision = prcurvesfiM2$curve[, 2])

pdf(file.path("M2AUCPRC.pdf"), width = 5, height = 5)
blineM2 <- (sum(dfM2$mortality == TRUE)) / length(dfM2$mortality)
blineM2
  
ggplot(dfprcsfiM2, aes(x = recall, y = precision)) +
  geom_line(color = "#27933E", size = 1) +
   geom_abline(intercept = blineM2, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = paste0("M2 SFI-10: Mortality (AUC = ", sprintf("%.3f", auprcsfiM2), ")"),
    x = "Sensitivity", y = "PPV") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))
plot(prcurvesfiM2)
dev.off()
```


# Testing models in R1

### Recode vars based on functional forms

```{r}
#######################
#age (age_R1)#
#######################
#check#
summary(dfR1$age_R1)

#no change#

#######################
#sex (sex_R1)#
#######################
#check#
table(dfR1$sex_R1)

dfR1$sex_R1 <- as.numeric(dfR1$sex_R1)

#no change#

#######################
#neighborhood clean?#
#(neighborhood_clean_R1)#
#######################
#check#
table(dfR1$neighborhood_clean_R1)

#code (1, 2) = (0), (3, 4) = (1)#
dfR1$neighborhood_clean_R1 <- ifelse(dfR1$neighborhood_clean_R1 == 1, 0, 
                                ifelse(dfR1$neighborhood_clean_R1 == 2, 0, 1))

#check#
table(dfR1$neighborhood_clean_R1)

#######################
#financial control#
#(financial_control_R1)#
#######################
#check#
table(dfR1$financial_control_R1)

#no change#

#######################
#family contact#
#(family_contact_R1)#
#######################
#check#
table(dfR1$family_contact_R1)

#code (<=7) to (0); (8) to (1)#
dfR1$family_contact_R1 <- ifelse(dfR1$family_contact_R1 == 8, 1, 0)

#check#
table(dfR1$family_contact_R1)

#######################
#friends contact#
#(friends_contact_R1)#
#######################
#check#
table(dfR1$friends_contact_R1)

#compress bottom (1, 2) = 1, subtract 1 from else#
dfR1$friends_contact_R1 <-  ifelse(dfR1$friends_contact_R1 <= 2, 1, 
                                              dfR1$friends_contact_R1 - 1)

#check#
table(dfR1$friends_contact_R1)

#######################
#work for pay?#
#(work_R1)#
#######################
#check#
table(dfR1$work_R1)

#no change#

#######################
#volunteer#
#(volunteer_R1)#
#######################
#check#
summary(dfR1$volunteer_R1)

#change to log scale#
dfR1$volunteer_R1 <- log(dfR1$volunteer_R1 + 1)

#check#
summary(dfR1$volunteer_R1)

#######################
#loneliness#
#(loneliness_R1)#
#######################
#check#
table(dfR1$loneliness_R1)

#no change#

#######################
#treated w/ less courtesy/respect#
#(courtresp_R1)#
#######################
#check#
table(dfR1$courtresp_R1)

#code (< 5) to (0); (>= 5) to (1)#
courtresp0 <- c(1, 2, 3, 4)
courtresp1 <- c(5, 6, 7)

dfR1$courtresp_R1 <- ifelse(dfR1$courtresp_R1 %in% courtresp0, 0, 1)

#check#
table(dfR1$courtresp_R1)
```

### Apply models

```{r}
#######################
#rename cols to be congruent#
#w HRS names#
#######################
colnames(dfR1)[c(5:11, 50:52)] <- c("sex", "age", "neighborhood", "financial_control", 
                             "freq_children", "activities_youth", "work", "volunteering", 
                             "feel_isolated", "courtresp")
```

### SFI10

```{r}
#######################
#specifying models#
#######################
dfR1$SFIlp <- predict(sfi, newdata = dfR1, type = "link")
dfR1$SFIpr <- predict(sfi, newdata = dfR1, type = "response")


#######################
#distribution viz#
#######################
pdf(file.path(outdir, "R1SFI10distributions.pdf"), width = 6, height = 5)
hist(dfR1$SFIlp, breaks = "sturges", freq = FALSE, 
     xlim = c(-8, 2), ylim = c(0, 0.3), 
     main = "R1 Cohort: SFI-10", xlab = "Mortality Risk (Log Probability)")
lines(density(dfR1$SFIlp), col = "#27933E")
hist(dfR1$SFIpr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 16), 
     main = "R1 Cohort: SFI-10", xlab = "Mortality Risk (Probability)")
lines(density(dfR1$SFIpr), col = "#27933E")
dev.off()
```

### ageglm

```{r}
#######################
#specifying models#
#######################
dfR1$agelp <- predict(glmage, newdata = dfR1, type = "link")
dfR1$agepr <- predict(glmage, newdata = dfR1, type = "response")

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "R1AGEdistributions.pdf"), width = 6, height = 5)
hist(dfR1$agelp, breaks = "sturges", freq = FALSE, 
     xlim = c(-6, -1), ylim = c(0, 0.4), 
     main = "R1 Cohort: Age", xlab = "Mortality Risk (Log Probability)")
lines(density(dfR1$agelp), col = "#27933E")
hist(dfR1$agepr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 20), 
     main = "R1 Cohort: Age", xlab = "Mortality Risk (Probability)")
lines(density(dfR1$agepr), col = "#27933E")
dev.off()
```

### age & sex

```{r}
#######################
#specifying models#
#######################
dfR1$sexagelp <- predict(sexage, newdata = dfR1, type = "link")
dfR1$sexagepr <- predict(sexage, newdata = dfR1, type = "response")

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "R1agesexdistributions.pdf"), width = 6, height = 5)
hist(dfR1$sexagelp, breaks = "sturges", freq = FALSE, 
     xlim = c(-6, -1), ylim = c(0, 0.4), 
     main = "R1 Cohort: Age & Sex", xlab = "Mortality Risk (Log Probability)")
lines(density(dfR1$sexagelp), col = "#27933E")
hist(dfR1$sexagepr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 25), 
     main = "R1 Cohort: Age & Sex", xlab = "Mortality Risk (Probability)")
lines(density(dfR1$sexagepr), col = "#27933E")
dev.off()
```

### SFI8

```{r}
#######################
#specifying models#
#######################
dfR1$socSFIlp <- predict(sfisoc, newdata = dfR1, type = "link")
dfR1$socSFIpr <- predict(sfisoc, newdata = dfR1, type = "response")

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "R1SFI8distributions.pdf"), width = 6, height = 5)
hist(dfR1$socSFIlp, breaks = "sturges", freq = FALSE, 
     xlim = c(-6, -1), ylim = c(0, 0.8), 
     main = "R1 Cohort: SFI-8", xlab = "Mortality Risk (Log Probability)")
lines(density(dfR1$socSFIlp), col = "#27933E")
hist(dfR1$socSFIpr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 25), 
     main = "R1 Cohort: SFI-8", xlab = "Mortality Risk (Probability)")
lines(density(dfR1$socSFIpr), col = "#27933E")
dev.off()
```

### R1 AUC - ROC

```{r}
#######################
#SFI-10#
#######################
set.seed(555)
roccurvesfiR1 <- roc(dfR1$mortality, dfR1$SFIpr, auc = TRUE, 
                     ci = TRUE, of = "auc", ci.method = "bootstrap", 
                     conf.level = 0.95, boot.n = 9999, boot.stratified = TRUE)
aucsfiR1 <- roccurvesfiR1$auc
ciaucsfiR1 <-  roccurvesfiR1$ci
print(aucsfiR1)
print(ciaucsfiR1)

dfbcarocsfiR1 <- data.frame(response = dfR1$mortality, 
                            predicted = dfR1$SFIpr)
bcaciaucsfiR1 <- stratboot(dfbcarocsfiR1, auroc, R = 9999, strata = dfbcarocsfiR1$response)
print(bcaciaucsfiR1)

dfrocsfiR1 <- data.frame(specificity = 1 - roccurvesfiR1$specificities, 
                    sensitivity = roccurvesfiR1$sensitivities)


pdf(file.path(outdir, "R1AUCROCs.pdf"), width = 4, height = 4)
ggplot(dfrocsfiR1, aes(x = specificity, y = sensitivity)) + 
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
    labs(title = paste0("SFI-10: Mortality (AUC = ", sprintf("%.3f",aucsfiR1), ")"), 
         x = "1 - Specificity", y = "Sensitivity") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))

#######################
#age#
#######################
roccurveageR1 <- roc(dfR1$mortality, dfR1$agepr, auc = TRUE, 
                     ci = TRUE, of = "auc", ci.method = "bootstrap", 
                     conf.level = 0.95, boot.n = 9999, boot.stratified = TRUE)
aucageR1 <- roccurveageR1$auc
ciaucageR1 <-  roccurveageR1$ci
print(aucageR1)
print(ciaucageR1)

dfbcarocageR1 <- data.frame(response = dfR1$mortality, 
                            predicted = dfR1$agepr)
bcaciaucageR1 <- stratboot(dfbcarocageR1, auroc, R = 9999, strata = dfbcarocageR1$response)
print(bcaciaucageR1)

dfrocageR1 <- data.frame(specificity = 1 - roccurveageR1$specificities, 
                    sensitivity = roccurveageR1$sensitivities)

ggplot(dfrocageR1, aes(x = specificity, y = sensitivity)) + 
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
    labs(title = paste0("Age: Mortality (AUC = ", sprintf("%.3f",aucageR1), ")"), 
         x = "1 - Specificity", y = "Sensitivity") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))

#######################
#age & sex#
#######################
roccurvesexageR1 <- roc(dfR1$mortality, dfR1$sexagepr, auc = TRUE, 
                     ci = TRUE, of = "auc", ci.method = "bootstrap", 
                     conf.level = 0.95, boot.n = 9999, boot.stratified = TRUE)
aucsexageR1 <- roccurvesexageR1$auc
ciaucsexageR1 <-  roccurvesexageR1$ci
print(aucsexageR1)
print(ciaucsexageR1)

dfbcarocsexageR1 <- data.frame(response = dfR1$mortality, 
                               predicted = dfR1$sexagepr)
bcaciaucsexageR1 <- stratboot(dfbcarocsexageR1, auroc, R = 9999, strata = dfbcarocsexageR1$response)
print(bcaciaucsexageR1)

dfrocsexageR1 <- data.frame(specificity = 1 - roccurvesexageR1$specificities, 
                    sensitivity = roccurvesexageR1$sensitivities)

ggplot(dfrocsexageR1, aes(x = specificity, y = sensitivity)) + 
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
    labs(title = paste0("Age & Sex: Mortality (AUC = ", sprintf("%.3f",aucsexageR1), ")"), 
         x = "1 - Specificity", y = "Sensitivity") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))

#######################
#SFI-8#
#######################
roccurvesfisocR1 <- roc(dfR1$mortality, dfR1$socSFIpr, auc = TRUE, 
                     ci = TRUE, of = "auc", ci.method = "bootstrap", 
                     conf.level = 0.95, boot.n = 9999, boot.stratified = TRUE)
aucsfisocR1 <- roccurvesfisocR1$auc
ciaucsfisocR1 <-  roccurvesfisocR1$ci
print(aucsfisocR1)
print(ciaucsfisocR1)

dfbcarocsfisocR1 <- data.frame(response = dfR1$mortality, 
                               predicted = dfR1$socSFIpr)
bcaciaucsfisocR1 <- stratboot(dfbcarocsfisocR1, auroc, R = 9999, strata = dfbcarocsfisocR1$response)
print(bcaciaucsfisocR1)

dfrocsfisocR1 <- data.frame(specificity = 1 - roccurvesfisocR1$specificities, 
                    sensitivity = roccurvesfisocR1$sensitivities)

ggplot(dfrocsfisocR1, aes(x = specificity, y = sensitivity)) + 
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
    labs(title = paste0("SFI-8: Mortality (AUC = ", sprintf("%.3f",aucsfisocR1), ")"), 
         x = "1 - Specificity", y = "Sensitivity") + 
  theme_minimal() +
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))


#######################
#combined plot#
#######################
dfrocsfiR1$model <- paste0("SFI-10 (", sprintf("%.3f", aucsfiR1), ")")
dfrocageR1$model <- paste0("Age (", sprintf("%.3f", aucageR1), ")")
dfrocsexageR1$model <- paste0("Age & Sex (", sprintf("%.3f", aucsexageR1), ")")
dfrocsfisocR1$model <- paste0("SFI-8 (", sprintf("%.3f", aucsfisocR1), ")")

dfrocsR1 <- rbind(dfrocsfiR1, dfrocageR1, dfrocsexageR1, dfrocsfisocR1)

dfrocsR1$model <- factor(dfrocsR1$model, levels = c(
  paste0("SFI-10 (", sprintf("%.3f", aucsfiR1), ")"),
  paste0("Age & Sex (", sprintf("%.3f", aucsexageR1), ")"),
  paste0("Age (", sprintf("%.3f", aucageR1), ")"),
  paste0("SFI-8 (", sprintf("%.3f", aucsfisocR1), ")")))

modcols <- setNames(c("#27933E", "#B8CDF8", "#6A4C93", "#E4CC37"),
                    c(paste0("SFI-10 (", sprintf("%.3f", aucsfiR1), ")"),
                      paste0("Age (", sprintf("%.3f", aucageR1), ")"),
                      paste0("Age & Sex (", sprintf("%.3f", aucsexageR1), ")"),
                      paste0("SFI-8 (", sprintf("%.3f", aucsfisocR1), ")")))


ggplot(dfrocsR1, aes(x = specificity, y = sensitivity, color = model)) +
  geom_line(size = 1) +
  scale_color_manual(values = modcols) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
  labs(title = "Figure 2A with legend",
       x = "1 - Specificity", y = "Sensitivity", color = expression("Model (AU"[ROC]*")")) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 11),
    axis.text = element_text(size = 11), legend.title = element_text(size = 11),
    legend.text = element_text(size = 11), 
    legend.position = "right")

ggplot(dfrocsR1, aes(x = specificity, y = sensitivity, color = model)) +
  geom_line(size = 1) +
  scale_color_manual(values = modcols) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
  labs(title = "Figure 2A",
       x = "1 - Specificity", y = "Sensitivity", color = expression("Model (AU"[ROC]*")")) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 11),
    axis.text = element_text(size = 11), legend.title = element_text(size = 11),
    legend.text = element_text(size = 11), 
    legend.position = "none")

dev.off()

#######################
#testing#
#######################
#BCa CI-based bootstrap test: SexAge vs. SFI-10#
dfdiffsexage <- data.frame(response = dfR1$mortality,
                           pred1 = dfR1$SFIpr,
                           pred2 = dfR1$sexagepr)

sfisexagediff <- stratboot(dfdiffsexage, aurocdiff, R = 9999, strata = dfdiffsexage$response, conf.level = 0.90)
print(sfisexagediff)

#BCa CI-based bootstrap test: SFI-8 vs. SFI-10#
dfdiffsfisoc <- data.frame(response = dfR1$mortality,
                           pred1 = dfR1$SFIpr,
                           pred2 = dfR1$socSFIpr)

sfisocdiff <- stratboot(dfdiffsfisoc, aurocdiff, R = 9999, strata = dfdiffsfisoc$response, conf.level = 0.90)
print(sfisocdiff)
```

### R1 AUC - PRC

```{r}
#######################
#baseline#
#######################
blineR1 <- (sum(dfR1$mortality == TRUE)) / length(dfR1$mortality)
blineR1

#######################
#SFI-10#
#######################
set.seed(555)
prcurvesfiR1 <- pr.curve(scores.class0 = dfR1$SFIpr, weights.class0 = dfR1$mortality, curve = TRUE)
auprcsfiR1 <- prcurvesfiR1$auc.integral
print(auprcsfiR1)

dfbcaprcsfiR1 <- data.frame(response = dfR1$mortality, 
                            predicted = dfR1$SFIpr)
bcaciauprcsfiR1 <- stratboot(dfbcarocsfiR1, auprc, R = 9999, strata = dfbcaprcsfiR1$response)
print(bcaciauprcsfiR1)


dfprcsfiR1 <- data.frame(recall = prcurvesfiR1$curve[, 1], 
                         precision = prcurvesfiR1$curve[, 2])

pdf(file.path(outdir, "R1AUCPRCs.pdf"), width = 4, height = 4)
ggplot(dfprcsfiR1, aes(x = recall, y = precision)) +
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = blineR1, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = paste0("SFI-10: Mortality (AUC = ", sprintf("%.3f", auprcsfiR1), ")"),
    x = "Sensitivity", y = "PPV") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))

plot(prcurvesfiR1)

#######################
#age#
#######################
prcurveageR1 <- pr.curve(scores.class0 = dfR1$agepr, weights.class0 = dfR1$mortality, curve = TRUE)
auprcageR1 <- prcurveageR1$auc.integral
print(auprcageR1)

dfbcaprcageR1 <- data.frame(response = dfR1$mortality, 
                            predicted = dfR1$agepr)
bcaciauprcageR1 <- stratboot(dfbcarocageR1, auprc, R = 9999, strata = dfbcaprcageR1$response)
print(bcaciauprcageR1)

dfprcageR1 <- data.frame(recall = prcurveageR1$curve[, 1], 
                         precision = prcurveageR1$curve[, 2])

ggplot(dfprcageR1, aes(x = recall, y = precision)) +
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = blineR1, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = paste0("Age: Mortality (AUC = ", sprintf("%.3f", auprcageR1), ")"),
    x = "Sensitivity", y = "PPV") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))

plot(prcurveageR1)

#######################
#age & sex#
#######################
prcurvesexageR1 <- pr.curve(scores.class0 = dfR1$sexagepr, weights.class0 = dfR1$mortality, curve = TRUE)
auprcsexageR1 <- prcurvesexageR1$auc.integral
print(auprcsexageR1)

dfbcaprcsexageR1 <- data.frame(response = dfR1$mortality, 
                               predicted = dfR1$sexagepr)
bcaciauprcsexageR1 <- stratboot(dfbcarocsexageR1, auprc, R = 9999, strata = dfbcaprcsexageR1$response)
print(bcaciauprcsexageR1)

dfprcsexageR1 <- data.frame(recall = prcurvesexageR1$curve[, 1], 
                         precision = prcurvesexageR1$curve[, 2])

ggplot(dfprcsexageR1, aes(x = recall, y = precision)) +
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = blineR1, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = paste0("Age & Sex: Mortality (AUC = ", sprintf("%.3f", auprcsexageR1), ")"),
    x = "Sensitivity", y = "PPV") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))

plot(prcurvesexageR1)

#######################
#SFI-8#
#######################
prcurvesocsfiR1 <- pr.curve(scores.class0 = dfR1$socSFIpr, weights.class0 = dfR1$mortality, curve = TRUE)
auprcsocsfiR1 <- prcurvesocsfiR1$auc.integral
print(auprcsocsfiR1)

dfbcaprcsocsfiR1 <- data.frame(response = dfR1$mortality, 
                               predicted = dfR1$socSFIpr)
bcaciauprcsocsfiR1 <- stratboot(dfbcaprcsocsfiR1, auprc, R = 9999, strata = dfbcaprcsocsfiR1$response)
print(bcaciauprcsocsfiR1)

dfprcsocsfiR1 <- data.frame(recall = prcurvesocsfiR1$curve[, 1], 
                         precision = prcurvesocsfiR1$curve[, 2])

ggplot(dfprcsocsfiR1, aes(x = recall, y = precision)) +
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = blineR1, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = paste0("SFI-8: Mortality (AUC = ", sprintf("%.3f", auprcsocsfiR1), ")"),
    x = "Sensitivity", y = "PPV") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))

plot(prcurvesocsfiR1)

#######################
#combined plot#
#######################
dfprcsfiR1$model <- paste0("SFI-10 (", sprintf("%.3f", auprcsfiR1), ")")
dfprcageR1$model <- paste0("Age (", sprintf("%.3f", auprcageR1), ")")
dfprcsexageR1$model <- paste0("Age & Sex (", sprintf("%.3f", auprcsexageR1), ")")
dfprcsocsfiR1$model <- paste0("SFI-8 (", sprintf("%.3f", auprcsocsfiR1), ")")

dfprcsR1 <- rbind(dfprcsfiR1, dfprcageR1, dfprcsexageR1, dfprcsocsfiR1)

dfprcsR1$model <- factor(dfprcsR1$model, levels = c(
  paste0("SFI-10 (", sprintf("%.3f", auprcsfiR1), ")"),
  paste0("Age & Sex (", sprintf("%.3f", auprcsexageR1), ")"),
  paste0("Age (", sprintf("%.3f", auprcageR1), ")"),
  paste0("SFI-8 (", sprintf("%.3f", auprcsocsfiR1), ")")))

modcols <- setNames(c("#27933E", "#B8CDF8", "#6A4C93", "#E4CC37"),
                    c(paste0("SFI-10 (", sprintf("%.3f", auprcsfiR1), ")"),
                      paste0("Age (", sprintf("%.3f", auprcageR1), ")"),
                      paste0("Age & Sex (", sprintf("%.3f", auprcsexageR1), ")"),
                      paste0("SFI-8 (", sprintf("%.3f", auprcsocsfiR1), ")")))


ggplot(dfprcsR1, aes(x = recall, y = precision, color = model)) +
  geom_line(size = 1) +
  scale_color_manual(values = modcols) +
  geom_abline(intercept = blineR1, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = "Supplemental Figure 2A with legend",
       x = "Sensitivity", y = "PPV", color = expression("Model (AU"[PRC]*")")) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 9),
    axis.text = element_text(size = 9), legend.title = element_text(size = 9),
    legend.text = element_text(size = 9), 
    legend.position = "right")

ggplot(dfprcsR1, aes(x = recall, y = precision, color = model)) +
  geom_line(size = 1) +
  scale_color_manual(values = modcols) +
  geom_abline(intercept = blineR1, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = "Supplemental Figure 2A",
       x = "Sensitivity", y = "PPV", color = expression("Model (AU"[PRC]*")")) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 9),
    axis.text = element_text(size = 9), legend.title = element_text(size = 9),
    legend.text = element_text(size = 9), 
    legend.position = "none")

dev.off()

#######################
#testing#
#######################
#BCa CI-based bootstrap test: SexAge vs. SFI-10#
sfisexageprcdiff <- stratboot(dfdiffsexage, auprcdiff, R = 9999, strata = dfdiffsexage$response, conf.level = 0.90)
print(sfisexageprcdiff)
```


# Feature importance functions

```{r}
aurocloss <- function(response, predicted) {
  rocobj <- roc(response, predicted)
  return(1 - rocobj$auc)
}

auprcloss <- function(response, predicted) {
  pc <- predicted[response == 1]
  nc <- predicted[response == 0]
  prcobj <- pr.curve(scores.class0 = pc, scores.class1 = nc)
  return(1 - prcobj$auc.integral)
}
```


# R1 Feature importance - ROC

```{r}
#######################
#SFI-10#
#######################
set.seed(555)
sfipredR1 <- Predictor$new(model = sfi, data = dfR1[, c(5:11, 50:52)], y = dfR1$mortality)

sfiimpR1 <- FeatureImp$new(sfipredR1, loss = aurocloss, compare = "difference", n.repetitions = 1000)

sfiimpplotR1 <- plot(sfiimpR1)

dfR1labs <- c(age = "Age", sex = "Sex", neighborhood = "Neighborhood clean", 
              financial_control = "Financial control", freq_children = "Family contact",
              activities_youth = "Friend contact", work = "Working", volunteering = "Volunteering", 
              feel_isolated = "Loneliness", courtresp = "Courtesy/respect")

pdf(file.path(outdir, "R1FeatureImpsROC.pdf"), width = 5.1, height = 4)
sfiimpplotR1 +
  ggtitle("Figure 2B") +
  scale_x_continuous(name = expression("Feature Importance (Loss: 1 - AU"[ROC]*")"), 
                     breaks = c(0.00, 0.05, 0.10, 0.15, 0.20), 
                     limits = c(-0.02, 0.20)) +
  scale_y_discrete(name = NULL, labels = dfR1labs) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13.5),      
    axis.text = element_text(size = 13.5))

sfiimpR1res <- sfiimpR1$results
sfiimpR1res
write.csv(sfiimpR1res, file.path(outdir, "MIDUS R1 SFI10 FeatureImpROC.csv"), row.names = TRUE)

#######################
#SFI-8#
#######################
socpredR1 <- Predictor$new(model = sfisoc, data = dfR1[, c(7:11, 50:52)], y = dfR1$mortality)

socimpR1 <- FeatureImp$new(socpredR1, loss = aurocloss, compare = "difference", n.repetitions = 1000)

socimpplotR1 <- plot(socimpR1)

socimpplotR1 +
  ggtitle("Figure 2C") +
  scale_x_continuous(name = expression("Feature Importance (Loss: 1 - AU"[ROC]*")"), 
                     breaks = c(0.00, 0.05, 0.10, 0.15, 0.20), 
                     limits = c(-0.02, 0.20)) +
  scale_y_discrete(name = NULL, labels = dfR1labs) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13.5),      
    axis.text = element_text(size = 13.5))

dev.off()

socimpR1res <- socimpR1$results
socimpR1res
write.csv(socimpR1res, file.path(outdir, "MIDUS R1 SFI8 FeatureImpROC.csv"), row.names = TRUE)
```


# R1 Feature importance - PRC

```{r}
#######################
#SFI-10#
#######################
set.seed(555)
sfiimpprcR1 <- FeatureImp$new(sfipredR1, loss = auprcloss, compare = "difference", n.repetitions = 1000)

sfiimpprcplotR1 <- plot(sfiimpprcR1)

pdf(file.path(outdir, "R1FeatureImpsPRC.pdf"), width = 5.1, height = 4)
sfiimpprcplotR1 +
  ggtitle("Supplemental Figure 2B") +
  scale_x_continuous(name = expression("Feature Importance (Loss: 1 - AU"[PRC]*")"), 
                     limits = c(-0.02, 0.10), 
                     breaks = c(0.00, 0.02, 0.04, 0.06, 0.08, 0.10)) +
  scale_y_discrete(name = NULL, labels = dfR1labs) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13.5),      
    axis.text = element_text(size = 13.5))

sfiimpprcR1res <- sfiimpprcR1$results
sfiimpprcR1res
write.csv(sfiimpprcR1res, file.path(outdir, "MIDUS R1 SFI10 FeatureImpPRC.csv"), row.names = TRUE)

#######################
#social predictors#
#######################
set.seed(555)
socimpprcR1 <- FeatureImp$new(socpredR1, loss = auprcloss, compare = "difference", n.repetitions = 1000)

socimpprcplotR1 <- plot(socimpprcR1)

socimpprcplotR1 +
  ggtitle("Supplemental Figure 2C") +
  scale_x_continuous(name = expression("Feature Importance (Loss: 1 - AU"[PRC]*")"), 
                     limits = c(-0.02, 0.06), 
                     breaks = c(0.00, 0.02, 0.04, 0.06)) + 
  scale_y_discrete(name = NULL, labels = dfR1labs) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13.5),      
    axis.text = element_text(size = 13.5))

dev.off()

socimpprcR1res <- socimpprcR1$results
socimpprcR1res
write.csv(socimpprcR1res, file.path(outdir, "MIDUS R1 SFI8 FeatureImpPRC.csv"), row.names = TRUE)
```


# Criterion testing

The following, in tandem, produce the results presented in Table 2.

### SFI-10 criterion

```{r}
######################
#only with CIs#
######################
mpqposR1 <- c("mpq_wellbeing_R1", "mpq_socpotency_R1", "mpq_achieve_R1", "mpq_socclose_R1")
mpqnegR1 <- c("mpq_stressreact_R1", "mpq_aggression_R1", "mpq_alienation_R1")

allvars <- c("health_R1", "mehealth_R1", "badl_R1", "iadl_R1", pwbR1, swbR1, "affpos_R1", 
             "affneg_R1", mpqposR1, mpqnegR1,
             "fmly_affectsolidarity_R1", "frnd_affectsolidarity_R1")


corrtablenop <- function(vars, targetvar, data) {
  results <- data.frame(
    variable = character(),
    mean_sd = character(),
    rho = character(),
    z = numeric(),
    stringsAsFactors = FALSE
  )
  
  for (var in vars) {
    spearman_corr <- cor(data[[targetvar]], data[[var]], method = "spearman")
    
    z <- 0.5 * log((1 + spearman_corr) / (1 - spearman_corr))
    
    sez <- 1 / sqrt(length(data[[targetvar]]) - 3)
    zlo <- z - qnorm(0.975) * sez
    zhi <- z + qnorm(0.975) * sez
    
    rholo <- (exp(2 * zlo) - 1) / (exp(2 * zlo) + 1)
    rhohi <- (exp(2 * zhi) - 1) / (exp(2 * zhi) + 1)
    
    rhocistr <- sprintf("%.2f [%.2f, %.2f]", spearman_corr, rholo, rhohi)
    rhocistr <- sub("^0\\.", ".", rhocistr)
    
    meanv <- mean(data[[var]], na.rm = TRUE)
    sdv <- sd(data[[var]], na.rm = TRUE)
    meansdstr <- sprintf("%.2f (%.2f)", meanv, sdv)
    
    results <- rbind(results, data.frame(
      variable = var,
      mean_sd = meansdstr,
      rho = rhocistr,
      z = z
    ))
  }
  
  #apa#
  results$z <- round(results$z, 2)
  
  return(results)
}

#save#
sfiallcors <- corrtablenop(allvars, "SFIlp", data = dfR1)
write.csv(sfiallcors, file.path(outdir, "R1sfiallcorrs.csv"), row.names = FALSE)
```

### SFI-8 criterion

```{r}
#corvars#
allvars <- c("health_R1", "mehealth_R1", "badl_R1", "iadl_R1", pwbR1, swbR1, 
             "affpos_R1", "affneg_R1", mpqposR1, mpqnegR1,
             "fmly_affectsolidarity_R1", "frnd_affectsolidarity_R1")

#save#
sfisocallcors <- corrtablenop(allvars, "socSFIlp", data = dfR1)
write.csv(sfisocallcors, file.path(outdir, "R1SFI8socallcorrs.csv"), row.names = FALSE)
```

### SFI-10 and SFI-8 different?

```{r}
lessvars <- c("health_R1", "mehealth_R1", pwbR1, swbR1, "affpos_R1", mpqposR1, 
              "fmly_affectsolidarity_R1", "frnd_affectsolidarity_R1")
greatervars <- c("affneg_R1", mpqnegR1, "badl_R1", "iadl_R1")

r_kh <- cor(dfR1$socSFIlp, dfR1$SFIlp, method = "spearman")
n <- nrow(dfR1)

comparecortable <- function(vars, data, alternative) {
  results <- data.frame(
    variable = character(),
    rho_socSFIlp = numeric(),
    rho_SFIlp = numeric(),
    p_value = numeric(),
    alternative = character(),
    stringsAsFactors = FALSE
  )
  
  for (var in vars) {
    r_jk <- cor(data[[var]], data$socSFIlp, method = "spearman")
    r_jh <- cor(data[[var]], data$SFIlp, method = "spearman")
    
    result <- cocor.dep.groups.overlap(
      r.jk = r_jk, 
      r.jh = r_jh, 
      r.kh = r_kh, 
      n = n, 
      alternative = alternative,
      alpha = 0.05,
      conf.level = 0.95
    )
    
    results <- rbind(results, data.frame(
      variable = var,
      rho_SFIlp = round(r_jh, 3),
      rho_socSFIlp = round(r_jk, 3),
      p_value = round(result@steiger1980$p.value, 4),
      alternative = alternative
    ))
  }
  
  return(results)
}

#less#
lesstestcors <- comparecortable(lessvars, dfR1, alternative = "less")

#greater#
greatertestcors <- comparecortable(greatervars, dfR1, alternative = "greater")

#combine
alltestcors <- rbind(lesstestcors, greatertestcors)

alltestcors$p_value_adj <- p.adjust(alltestcors$p_value, method = "BH")

print(alltestcors)
write.csv(alltestcors, file.path(outdir, "R1sfitestcorss.csv"), row.names = FALSE)
```


### Age + criterion checks

```{r}
#corvars#
allvars <- c("health_R1", "mehealth_R1", "badl_R1", "iadl_R1", pwbR1, swbR1, 
             "affpos_R1", "affneg_R1", mpqposR1, mpqnegR1,
             "fmly_affectsolidarity_R1", "frnd_affectsolidarity_R1")

#save#
ageallcors <- corrtablenop(allvars, "age", data = dfR1)
write.csv(ageallcors, file.path(outdir, "R1ageallcorrs.csv"), row.names = FALSE)
```


# Correlations between age + SFI vars

```{r}
#######################
#need to recode variables first so that higher value#
#direction associated with mortality in SFI#
#######################
colnames(dfR1oc)[c(5:11, 50:52)] <- c("sex", "age", "neighborhood", "financial_control", 
                                      "freq_children", "activities_youth", "work", "volunteering", 
                                      "feel_isolated", "courtresp")
#age#
#no change#

#sex#
dfR1oc$sex <- as.numeric(dfR1oc$sex)

#neighborhood clean?#
#no change#
table(dfR1oc$neighborhood)

#financial control#
#REVERSE 0-10#
table(dfR1oc$financial_control)
dfR1oc$financial_control <- 10 - dfR1oc$financial_control
table(dfR1oc$financial_control)

#family contact#
#no change#
table(dfR1oc$freq_children)

#friend contact#
#no change#
table(dfR1oc$activities_youth)

#work#
#REVERSE 0-1#
table(dfR1oc$work)
dfR1oc$work <- 1 - dfR1oc$work
table(dfR1oc$work)

#volunteer#
#no change#

#feel isolated#
#currently 1 = high, 13 = low#
#REVERSE 1-13#
table(dfR1oc$feel_isolated)
dfR1oc$feel_isolated <- 14 - dfR1oc$feel_isolated
table(dfR1oc$feel_isolated)

#treated courtesy/respect#
table(dfR1oc$courtresp)
dfR1oc$courtresp <- 8 - dfR1oc$courtresp
table(dfR1oc$courtresp)

#######################
#correlations of social items with age raw HRS#
#######################
ogcordfspearR1 <- dfR1oc[c(6, 7:10, 51, 52)]
view(ogcordfspearR1)
ogcordfpearR1 <- dfR1oc[c(6, 11, 50)]
view(ogcordfpearR1)

ogcordatpearR1 <- cor(ogcordfpearR1, method = "pearson")
ogcordatspearR1 <- cor(ogcordfspearR1, method = "spearman")
subsetogpearR1 <- ogcordatpearR1["age", ,drop = FALSE]
subsetogspearR1 <- ogcordatspearR1["age", ,drop = FALSE]

ogcordR1 <- cbind(subsetogpearR1, subsetogspearR1)
ogcordatR1 <- as.data.frame(ogcordR1)
ogcordatR1 <- ogcordatR1[, c("age", "age", "freq_children", "activities_youth", "volunteering", 
                          "feel_isolated", "neighborhood", "financial_control", "courtresp", 
                          "work")]
ogcordatR1 <- ogcordatR1[, c(-2)]
ogcordatR1 <- as.matrix(ogcordatR1)
colnames(ogcordatR1) <- dfR1labs[colnames(ogcordatR1)]
rownames(ogcordatR1) <- dfR1labs[rownames(ogcordatR1)]

#visualize#
png(file.path(outdir, "R1agesocialitemscors.png"), width = 9, height = 9, units = "in", res = 600)
socogplot <- corrplot(t(ogcordatR1), method = "color", addCoef.col = "#000000", 
                      col = colorRampPalette(c("#6A4C93", "white", "#27933E"))(200),  
                      col.lim = c(-1, 1), tl.cex = 1.4, tl.col = "grey30", 
                      cl.pos = 'n', number.cex = 1)
dev.off()

#######################
#following up on positive correlations#
#######################
#volunteering#
cor.test(dfR1oc$age, dfR1oc$volunteering, alternative = "greater", 
         method = "pearson", exact = NULL, conf.level = 0.95)

#working
cor.test(dfR1oc$age, dfR1oc$work, alternative = "greater", 
         method = "pearson", exact = NULL, conf.level = 0.95)
```
