---
title: "HRS SFI Analysis"
author: "Charlie Collinge"
date: "2026-01-09"
output: html_document
---

# Set up 

```{r}
knitr::opts_chunk$set(echo = TRUE)

#setup#
library(tidyverse)
library(haven)
library(openxlsx)
library(corrplot)
library(RColorBrewer)
library(pROC)
library(PRROC)
library(iml)
library(rms)
library(boot)

#specify an out directory for results & visualization#
outdir <- "/example/directory/"

#set working directory to location of data files#
setwd("/example/HRSfiles/")

#import RAND HRS longitudinal file 2020 V2 released May 2024, downloaded 3/17/25#
lrand <- read_sav(paste0(getwd(), "/randhrs1992_2020v2_SPSS/randhrs1992_2020v2.sav"))
names(lrand) <- tolower(names(lrand))
```


# Completion of Form LB in 2010 & 65+ 

```{r}
#Shah et al. use the HRS cross-wave tracker (2016 version) to ascertain#
#their analytic cohorts, and then combine this with RAND HRS fat files for further analysis.#
#See: https://github.com/sachinjshah/Social-Frailty-Index/blob/main/Step1_cohort_identification.sas#

#RAND HRS data products were constructed using the HRS core data products (e.g. the tracker file)#
#and contain cleaned and processed variables with consistent and intuitive naming conventions,# 
#model-based imputations, and spousal counterparts of most individual-level variables.#
#See: https://www.rand.org/health/surveys/hrs.html. 

#To ensure consistency in variable naming and imputation practices, only RAND HRS products#
#were used in this analysis. Thus Shah et al.'s cohort identification pipeline was conducted#
#substituting analogous RAND variables where necessary.#

#Only include individuals who completed any#
#leave behind questionnaire between 2006 and 2014 (waves 8-12).#
#This is a replication of Shah et al.'s cohort identification procedure#
#but is not necessarily pertinent here since final cohort specifications#
#were already published.#
lbcols <- c("r8lbcomp", "r9lbcomp", "r10lbcomp", "r11lbcomp", "r12lbcomp")
lrand <- lrand[apply(!is.na(lrand[, lbcols]), 1, any), ]

#2010 completion#
c10 <- lrand[lrand$r10lbcomp == 1, ]
c10$lb_year <- 2010
c10 <- c10[!is.na(c10$r10lbcomp), ]

#N = 8332 (8332 reported at https://github.com/sachinjshah/Social-Frailty-Index); 

#age 65+?#
#note: Shah et al. use "mage > 65", so technically 66+#
#using r10agem_e for age (in months to use most restrictive unit)# 
#at END of interview (see RAND longitudinal p. 129).#
#Note: 66 * 12 = 792 so "> 791"#
c10 <- c10[c10$r10agem_e > 791, ]

#N = 4308 (4302 reported by Shah et al.)#
```


# Create final df for 2010 cohort 

```{r}
#set wd#
setwd("/example/HRSfiles/")

#required for sample descriptive stats#
lranddescvars <- c("r10iwstat", "hacohort", 
                   "raracem", "rahispan", "raeduc", 
                   "r10shlt", "r10diabe", "r10cancre",  "r10hearte",
                   "r10stroke", "r10arthre", 
                   "r10smokev", "r10smoken",
                   "h10itot")

#required for sfi#
lrandsfivars <- c("hhidpn", "r10lbcomp", "lb_year", "ragender", "r10agey_e", 
               "r10agem_e", "r10iwbeg", "r10iwendm", "r10iwendy", 
               "radyear", "radmonth")

c10 <- c10[, c(lranddescvars, lrandsfivars)]

#convert r10iwbeg from SAS f8.0 date (days since 1/1/1960)#
#to year and month#
c10$r10iwbeg_r <- as.Date(c10$r10iwbeg, origin = "1960-01-01")

c10$iwbeg_y <- format(c10$r10iwbeg_r, "%Y")
c10$iwbeg_y <- as.numeric(c10$iwbeg_y)
c10$iwbeg_m <- format(c10$r10iwbeg_r, "%m")
c10$iwbeg_m <- as.numeric(c10$iwbeg_m)

#import RAND HRS 2010 fat file V6A released July 2022, downloaded 3/17/25#
rand10 <- read_sav(paste0(getwd(), "/hd10f6a_SPSS/hd10f6a.sav"))
names(rand10) <- tolower(names(rand10))

clb10 <- merge(c10, rand10, by = "hhidpn", all.x = TRUE)

df10 <- clb10[, c(lranddescvars, lrandsfivars, "iwbeg_y", "iwbeg_m",
                  "md171", "mc095", "mc103", "mc104", "mc105",
                  "mlb021f", "mlb026", "mlb009a", "mlb001b", 
                  "mlb045", "mlb001d", "mlb020c", "mlb030a")]
```


# Mortality 

```{r}
#basing "4yr" mortality calculation on Shah et al.'s actual criteria (not 4yr mortality)#
#via authors: We mentioned "4-year mortality" in the paper, but the way we defined the#
#death_flag (the outcome, a binary indicator) was to include all deaths that occurred by#
#the end of 2015 (Dec 31, 2015) for the 2010 cohort.#

#note: dates cannot be narrowed down from the month level#
#note: <= 2016 because December, 2015 would be 2015 + 1#
#in the formula below, so 2016#

df10$DOD <- df10$radyear + df10$radmonth / 12
df10$mortality <- ifelse(is.na(df10$DOD), 0,
                         ifelse(df10$DOD <= 2016, 1, 0))

table(df10$mortality == 1)
#935/4308 compared to 960/4302 published#

#checking when died#
df10$interview <- df10$r10iwendy + df10$r10iwendm / 12
df10$mortalitytime <- df10$DOD - df10$interview

summary(df10$mortalitytime[df10$mortality == 1])
hist(df10$mortalitytime[df10$mortality == 1], breaks = "sturges", freq = FALSE, 
     main = "2010 Cohort: Mortality Time", xlab = "Time (Years)")
lines(density(df10$mortalitytime[df10$mortality == 1]), col = "#27933E")

#checking outcome assessment window#
df10$outcometime <- 2016 - df10$interview
summary(df10$outcometime)
```


Total mortality suggested = 935
Max mortality time = 5.75 years
Shah et al. (2023) PNAS reported = 960


# 2010 Descriptive Stats 

The following 2 sections can be combined to recreate the descriptive statistics in Table S1.

### Descriptive stats - demographics, health status, function 

```{r}
#####################
#demographics#
#####################
#age (median+IQR)#
summary(df10$r10agey_e)
hist(df10$r10agey_e, breaks = "sturges", freq = FALSE, 
     main = "2010 Cohort: Age", xlab = "Age (Years)")
lines(density(df10$r10agey_e), col = "#27933E")

#sex (female+%)#
table(df10$ragender == 2)


#####################
#health status#
#####################
#fair or poor self reported health (N+%)#
table(!is.na(df10$r10shlt))
table(df10$r10shlt  %in% c(4, 5))

#diabetes (N+%)#
table(df10$r10diabe == 1)

#cancer (excluding minor skin cancer) (N+%)#
table(df10$r10cancre == 1)

#heart disease (N+%)#
table(df10$r10hearte == 1)

#stroke (N+%)#
table(df10$r10stroke == 1)

#arthritis (N+%)#
table(df10$r10arthre == 1)

#current smoker? (N+%)#
table(df10$r10smoken == 1)

#former smoker? (N+%)#
table(df10$r10smokev == 1 & df10$r10smoken == 0)


#####################
#mortality#
#####################
#mortality#
table(df10$mortality == 1)


#####################
#summarize#
#####################
N <- 4308

freq_percent <- function(var, condition) {
  freq <- sum(var %in% condition, na.rm = TRUE)  
  perc <- (freq / N) * 100              
  return(paste0(freq, " (", round(perc, 2), "%)"))
}

freq_percentlog <- function(var, condition) {
  freq <- sum(condition, na.rm = TRUE)
  perc <- (freq / length(var)) * 100
  paste0(freq, " (", round(perc, 2), "%)")
}

# Generate table with relevant variables
c10stats <- data.frame(
  Variable = c(
    "Fair or Poor Self-Reported Health",
    "Diabetes",
    "Cancer (excluding minor skin cancer)",
    "Heart Disease",
    "Stroke",
    "Arthritis",
    "Current Smoker",
    "Former Smoker"
  ),
  `N (%)` = c(
    freq_percent(df10$r10shlt, c(4, 5)), 
    freq_percent(df10$r10diabe, 1),  
    freq_percent(df10$r10cancre, 1),  
    freq_percent(df10$r10hearte, 1),  
    freq_percent(df10$r10stroke, 1),  
    freq_percent(df10$r10arthre, 1),  
    freq_percent(df10$r10smoken, 1),  
    freq_percent(df10$r10smokev[df10$r10smoken == 0], 1)
  )
)

write.csv(c10stats, file.path(outdir, "2010 HRS health and function.csv"), row.names = TRUE)
```


### Descriptive stats - social characteristics 

```{r}
#####################
#general#
#####################
#race (N+%)#
table(df10$raracem)
table(!is.na(df10$raracem))

#ethnicity (N+%)#
table(df10$rahispan)
table(!is.na(df10$rahispan))

#education (N+%)#
table(df10$raeduc)
table(!is.na(df10$raeduc))

#income (N+%)#
summary(df10$h10itot)
hist(df10$h10itot, breaks = "sturges", freq = FALSE, 
     xaxt = 'n', ylim = c(0, 0.000020),
     main = "2010 Cohort: Household Income", xlab = "Annual Income (USD)")
axis(side = 1, at = pretty(range(df10$h10itot)), labels = pretty(range(df10$h10itot)))
lines(density(df10$h10itot), col = "#27933E")


#####################
#SFI-10 vars#
#####################
#all tables based on functional forms analysis below#
#neighborhood unclean#
#code (< 4) to (0); (>= 4) to (1)#
table(df10$mlb021f >= 4)

#low financial control#
#code (< 2) to (0); (>= 2) to (1)#
table(df10$mlb026 < 2)

#meets with children less than once/year#
#code (<= 5) to (0); (6) to (1)#
table(df10$mlb009a == 6)

#activities w/ youth#
summary(df10$mlb001b)

#work#
table(df10$mlb045 == 1)

#no volunteer#
#code (<= 6) to (0); (7) to (1)#
table(df10$mlb001d == 7)

#loneliness#
table(df10$mlb020c)

#courtesy/respect#
#code (<= 4) to (0); (5, 6) to (1)#
table(df10$mlb030a <= 4)

#####################
#summarize#
#####################
N = 4308

c10descstats <- data.frame(
  Variable = c(
    "Black",
    "White",
    "Other", 
    "Hispanic origin", 
    "Not of hispanic origin", 
    "Less than high school", 
    "High school diploma or equiv", 
    "Some college", 
    "College and above",
    "< $56,599", 
    "$56,600 - 169,799", 
    "> $169,800", 
    "Age", 
    "Sex (Female)", 
    "Neighborhood is not clean", 
    "Low percieved financial control", 
    "Meets with children less than once per year or equivalent", 
    "Activities with younger folks or equivalent", 
    "Presently working", 
    "Does no volunteer/charity work", 
    "Feels isolated often", 
    "Feels isolated some of the time", 
    "Feels isolated hardly ever or never",
    "Often treated with less courtesy and respect than others"
  ),
  `N (%)` = c(
    freq_percent(df10$raracem, 2),
    freq_percent(df10$raracem, 1),
    freq_percent(df10$raracem, 3),
    freq_percent(df10$rahispan, 1), 
    freq_percent(df10$rahispan, 0),
    freq_percent(df10$raeduc, 1),
    freq_percent(df10$raeduc, 2:3),
    freq_percent(df10$raeduc, 4),
    freq_percent(df10$raeduc, 5),
    freq_percentlog(df10$h10itot, df10$h10itot < 56600),
    freq_percentlog(df10$h10itot, df10$h10itot >= 56600 & df10$h10itot <= 169799),
    freq_percentlog(df10$h10itot, df10$h10itot > 169799),
    paste0(round(median(df10$r10agey_e), 2), 
           " [", round(summary(df10$r10agey_e)[2], 2),",", 
           round(summary(df10$r10agey_e)[5], 2), "]"), 
    freq_percent(df10$ragender, 2), 
    freq_percent(df10$mlb021f, 4:10), 
    freq_percent(df10$mlb026, 0:1), 
    freq_percent(df10$mlb009a, 6), 
    paste0(round(median(df10$mlb001b), 2), 
           " [", round(summary(df10$mlb001b)[2], 2), ",",
           round(summary(df10$mlb001b)[5], 2), "]"),
    freq_percent(df10$mlb045, 1), 
    freq_percent(df10$mlb001d, 7), 
    freq_percent(df10$mlb020c, 1), 
    freq_percent(df10$mlb020c, 2),
    freq_percent(df10$mlb020c, 3),
    freq_percent(df10$mlb030a, 1:4)
  )
)

write.csv(c10descstats, file.path(outdir, "2010 HRS descriptive stats.csv"), row.names = TRUE)
```


# Find median (continuous) & mode values 

```{r}
findmode <- function(x) {
  freq_table <- table(x)
  mode <- as.numeric(names(freq_table[freq_table == max(freq_table)]))
  return(mode)
}

#######################
#age (r10agey_e)#
#######################
median(df10$r10agey_e, na.rm = TRUE)

#median: 75#

#######################
#gender (ragender)#
#######################
findmode(df10$ragender)

#mode: 2 (woman)#

#######################
#neighborhood clean (mlb021f)#
#######################
findmode(df10$mlb021f)

#mode: 1 (highest agreement with "this area is kept very clean")#

#######################
#financial control (mlb026)#
#######################
findmode(df10$mlb026)

#mode: 10 (very much control)#

#######################
#has children/freq. seeing (mlb009a)#
#######################
findmode(df10$mlb009a)

#mode: 2 (once or twice a week)#

#######################
#activities w/ younger folks; grandchildren, etc. (mlb001b)#
#######################
findmode(df10$mlb001b)

#mode: 7 (never/not relevant) *note; Shah et al. imputed "5" as the mode value#
#(https://github.com/sachinjshah/Social-Frailty-Index/blob/main/Detailed%20instructions#
#%20to%20calculate%20SFI%20with%20user%20input.pdf) did they treat this as continuous?#
median(df10$mlb001b, na.rm = TRUE)

#median: 5 (at least once a month)#

#######################
#work (mlb045)#
#######################
findmode(df10$mlb045)

#mode: 5 (No)#

#######################
#other volunteer/charity work (mlb001d)#
#######################
findmode(df10$mlb001d)

#mode: 7 (never/not relevant)#

#######################
#feel isolated (mlb020c)#
#######################
findmode(df10$mlb020c)

#mode: 3 (hardly ever or never)#

#######################
#treated w/ less courtesy/respect (mlb030a)#
#######################
findmode(df10$mlb030a)

#mode: 6 (never)#
```


# Single imputation of median and mode for missing values 

```{r}
#######################
#age (r10agey_e)#
#######################
table(!is.na(df10$r10agey_e))
df10$r10agey_e[is.na(df10$r10agey_e)] <- 75
table(!is.na(df10$r10agey_e))

#######################
#sex (ragender)#
#######################
table(!is.na(df10$ragender))
df10$ragender[is.na(df10$ragender)] <- 2
table(!is.na(df10$ragender))

#######################
#neighborhood clean (mlb021f)#
#######################
table(!is.na(df10$mlb021f))
df10$mlb021f[is.na(df10$mlb021f)] <- 1
table(!is.na(df10$mlb021f))

#######################
#financial control (mlb026)#
#######################
table(!is.na(df10$mlb026))
df10$mlb026[is.na(df10$mlb026)] <- 10
table(!is.na(df10$mlb026))

#######################
#has children/freq. seeing (mlb009a)#
#######################
table(!is.na(df10$mlb009a))
df10$mlb009a[is.na(df10$mlb009a)] <- 2
table(!is.na(df10$mlb009a))

#######################
#activities w/ younger folks; grandchildren, etc. (mlb001b)#
#######################
#creating dummy mlb001b var for comparison to what Shah et al. (2023) did#
df10$mlb001bdummy <- df10$mlb001b
table(!is.na(df10$mlb001bdummy))
df10$mlb001bdummy[is.na(df10$mlb001bdummy)] <- 5
table(!is.na(df10$mlb001bdummy))

#true recoding#
table(!is.na(df10$mlb001b))
df10$mlb001b[is.na(df10$mlb001b)] <- 7
table(!is.na(df10$mlb001b))

#######################
#work (mlb045)#
#######################
table(!is.na(df10$mlb045))
df10$mlb045[is.na(df10$mlb045)] <- 5
table(!is.na(df10$mlb045))

#######################
#other volunteer/charity work (mlb001d)#
#######################
table(!is.na(df10$mlb001d))
df10$mlb001d[is.na(df10$mlb001d)] <- 7
table(!is.na(df10$mlb001d))

#######################
#feel isolated (mlb020c)#
#######################
table(!is.na(df10$mlb020c))
df10$mlb020c[is.na(df10$mlb020c)] <- 3
table(!is.na(df10$mlb020c))

#######################
#treated w/ less courtesy/respect (mlb030a)#
#######################
table(!is.na(df10$mlb030a))
df10$mlb030a[is.na(df10$mlb030a)] <- 6
table(!is.na(df10$mlb030a))
```


# Recode vars to manual PNAS coding 

```{r}
#######################
#create dummy df for og coding#
#######################
df10og <- df10
df10ff <- df10

#######################
#age (r10agey_e)#
#no change in Shah et al. (2023)#
#######################

#######################
#gender (ragender)#
#######################
#check#
table(df10ff$ragender)

#code male (1) to (0); female (2) to (1) per Shah et al. (2023)#
df10ff$ragender[df10ff$ragender == 1] <- 0
df10ff$ragender[df10ff$ragender == 2] <- 1

#check#
table(df10ff$ragender)

#######################
#neighborhood clean? (mlb021f)#
#######################
#check#
table(df10ff$mlb021f)

#code (< 4) to (0); (>= 4) to (1) per Shah et al. (2023)#
df10ff$mlb021f <- ifelse(df10ff$mlb021f < 4, 0, 1)

#check#
table(df10ff$mlb021f)

#######################
#financial control (mlb026)#
#######################
#check#
table(df10ff$mlb026)

#code (< 4) to (0); (>= 4) to (1) per Shah et al. (2023)#
#BUT, since categorical model actually preferred, used#
#(< 2) to (0) and >= 2 to (1) to increase interpretability#
#and got lowest BIC#
df10ff$mlb026 <- ifelse(df10ff$mlb026 < 2, 0, 1)

#check#
table(df10ff$mlb026)

#######################
#has children/freq. seeing (mlb009a)#
#######################
#check#
table(df10ff$mlb009a)

#code (<= 5) to (0); (6) to (1) per Shah et al. (2023)#
df10ff$mlb009a <- ifelse(df10ff$mlb009a < 6, 0, 1)

#check#
table(df10ff$mlb009a)

#######################
#activities w/ younger folks; grandchildren, etc. (mlb001b)#
#no change in Shah et al. (2023)
#######################
#check#
table(df10ff$mlb001b)

#######################
#work (mlb045)#
#######################
#check#
table(df10ff$mlb045)

#code (5) to (0), no change in (1) per Shah et al. (2023)#
df10ff$mlb045[df10ff$mlb045 == 5] <- 0
  
#check#
table(df10ff$mlb045)

#######################
#other volunteer/charity work (mlb001d)#
#######################
#check#
table(df10ff$mlb001d)

#code (<= 6) to (0); (7) to (1) per Shah et al. (2023)#
df10ff$mlb001d <- ifelse(df10ff$mlb001d < 7, 0, 1)

#check#
table(df10ff$mlb001d)

#######################
#feel isolated (mlb020c)#
#######################
#check#
table(df10ff$mlb020c)

#code (1) to (0); (2) to (1); (3) to (2) per Shah et al. (2023)#
df10ff$mlb020c <- df10ff$mlb020c - 1

#check#
table(df10ff$mlb020c)

#######################
#treated w/ less courtesy/respect (mlb030a)#
#######################
#check#
table(df10ff$mlb030a)

#code (<= 4) to (0); (5, 6) to (1) per Shah et al. (2023)# 
df10ff$mlb030a <- ifelse(df10ff$mlb030a < 5, 0, 1)

#check#
table(df10ff$mlb030a)
```


# Functional forms 

```{r}
#######################
#var types#
#######################
continuousvars <- c("age (r10agey_e)")
ordinalvars <- c("neighborhood clean (mlb021f)", "financial control (mlb026)", 
                 "has children/freq. seeing (mlb009a)","activities w/ younger folks; grandchildren, etc. (mlb001b)", 
                 "other volunteer/charity work (mlb001d)", 
                 "feel isolated (mlb020c)", "treated w/ less courtesy/respect (mlb030a)")
binaryvars <- c("gender (gender)", "currently working (mlb045)")

#note: not assessing functional forms of binary predictors#

#######################
#continuous#
#######################

########
#age#
alin <- glm(mortality ~ r10agey_e, data = df10og, family = binomial, na.action = na.omit)

df10og$logage <- log(df10og$r10agey_e + 1)
alog <- glm(mortality ~ logage, data = df10og, family = binomial, na.action = na.omit)

df10og$expage <- exp(df10og$r10agey_e)
aexp <- glm(mortality ~ expage, data = df10og, family = binomial, na.action = na.omit)

abic <- data.frame(
  Model = c("Linear", "Log", "Exp"),
  BIC = c(BIC(alin), BIC(alog), BIC(aexp))
)

print(abic)


#######################
#ordinal#
#######################

########
#neighborhood clean (mlb021f)#
nclin <- glm(mortality ~ mlb021f, data = df10og, family = binomial, na.action = na.omit)

df10og[, "mlb021f"] <- as.factor(df10og[,"mlb021f"])
nccat <- glm(mortality ~ mlb021f, data = df10og, family = binomial, na.action = na.omit)

ncman <- glm(mortality ~ as.factor(mlb021f), data = df10ff, family = binomial, na.action = na.omit)

ncbic <- data.frame(
  Model = c("Linear", "Categorical", "Binary"),
  BIC = c(BIC(nclin), BIC(nccat), BIC(ncman))
)

print(ncbic)

########
#financial control (mlb026)# 
fclin <- glm(mortality ~ mlb026, data = df10og, family = binomial, na.action = na.omit)

df10og[, "mlb026"] <- as.factor(df10og[,"mlb026"])
fccat <- glm(mortality ~ mlb026, data = df10og, family = binomial, na.action = na.omit)

fcman <- glm(mortality ~ as.factor(mlb026), data = df10ff, family = binomial, na.action = na.omit)

fcbic <- data.frame(
  Model = c("Linear", "Categorical", "Binary"),
  BIC = c(BIC(fclin), BIC(fccat), BIC(fcman))
)

print(fcbic)


########
#has children/freq. seeing (mlb009a)# 
hcfslin <- glm(mortality ~ mlb009a, data = df10og, family = binomial, na.action = na.omit)

df10og[, "mlb009a"] <- as.factor(df10og[,"mlb009a"])
hcfscat <- glm(mortality ~ mlb009a, data = df10og, family = binomial, na.action = na.omit)

hcfsman <- glm(mortality ~ as.factor(mlb009a), data = df10ff, family = binomial, na.action = na.omit)

hcfsbic <- data.frame(
  Model = c("Linear", "Categorical", "Binary"),
  BIC = c(BIC(hcfslin), BIC(hcfscat), BIC(hcfsman))
)

print(hcfsbic)


########
#activities w/ younger folks; grandchildren, etc. (mlb001b)#
ayflin <- glm(mortality ~ mlb001b, data = df10og, family = binomial, na.action = na.omit)

df10og[, "mlb001b"] <- as.factor(df10og[,"mlb001b"])
ayfcat <- glm(mortality ~ mlb001b, data = df10og, family = binomial, na.action = na.omit)

ayfman <- glm(mortality ~ as.factor(mlb001b), data = df10ff, family = binomial, na.action = na.omit)

ayfbic <- data.frame(
  Model = c("Linear", "Categorical", "Binary"),
  BIC = c(BIC(ayflin), BIC(ayfcat), BIC(ayfman))
)

print(ayfbic)


########
#other volunteer/charity work (mlb001d)#
vlin <- glm(mortality ~ mlb001d, data = df10og, family = binomial, na.action = na.omit)

df10og[, "mlb001d"] <- as.factor(df10og[,"mlb001d"])
vcat <- glm(mortality ~ mlb001d, data = df10og, family = binomial, na.action = na.omit)

vman <- glm(mortality ~ as.factor(mlb001d), data = df10ff, family = binomial, na.action = na.omit)

vbic <- data.frame(
  Model = c("Linear", "Categorical", "Binary"),
  BIC = c(BIC(vlin), BIC(vcat), BIC(vman))
)

print(vbic)

########
#feel isolated (mlb020c)#
lolin <- glm(mortality ~ mlb020c, data = df10og, family = binomial, na.action = na.omit)

df10og[, "mlb020c"] <- as.factor(df10og[,"mlb020c"])
locat <- glm(mortality ~ mlb020c, data = df10og, family = binomial, na.action = na.omit)

loman <- glm(mortality ~ as.factor(mlb020c), data = df10ff, family = binomial, na.action = na.omit)

lobic <- data.frame(
  Model = c("Linear", "Categorical", "Binary"),
  BIC = c(BIC(lolin), BIC(locat), BIC(loman))
)

print(lobic)


########
#treated w/ less courtesy/respect (mlb030a)#
crlin <- glm(mortality ~ mlb030a, data = df10og, family = binomial, na.action = na.omit)

df10og[, "mlb030a"] <- as.factor(df10og[,"mlb030a"])
crcat <- glm(mortality ~ mlb030a, data = df10og, family = binomial, na.action = na.omit)

crman <- glm(mortality ~ as.factor(mlb030a), data = df10ff, family = binomial, na.action = na.omit)

crbic <- data.frame(
  Model = c("Linear", "Categorical", "Binary"),
  BIC = c(BIC(crlin), BIC(crcat), BIC(crman))
)

print(crbic)

summary <- cbind(abic$BIC, ncbic$BIC, fcbic$BIC, hcfsbic$BIC, ayfbic$BIC,
                 vbic$BIC, lobic$BIC, crbic$BIC)
colnames(summary) <- c("age", "neighborhood cleanliness", "financial control", 
                       "freq. see children", "activities with youth", "volunteer", "feel isolated", "courtesy/respect")
rownames(summary) <- c("1", "2", "3")
summary <- as.data.frame(summary)
summary

write.csv(summary, file.path(outdir, "HRS functional forms.csv"), row.names = FALSE)
```


# Recode variables based on functional forms 

```{r}
#######################
#age (r10agey_e)#
#no change#
#######################

#######################
#gender (gender)#
#######################
#check#
table(df10$ragender)

#code male (1) to (0); female (2) to (1) per Shah et al. (2023)#
df10$ragender[df10$ragender == 1] <- 0
df10$ragender[df10$ragender == 2] <- 1

#check#
table(df10$ragender)

#######################
#neighborhood clean? (mlb021f)#
#######################
#check#
table(df10$mlb021f)

#code (< 4) to (0); (>= 4) to (1) per Shah et al. (2023)#
df10$mlb021f <- ifelse(df10$mlb021f < 4, 0, 1)

#check#
table(df10$mlb021f)

#######################
#financial control (mlb026)#
#######################
#check#
table(df10$mlb026)

#code (< 2) to (0); (>= 2) to (1) per my manual categorization#
df10$mlb026 <- ifelse(df10$mlb026 < 2, 0, 1)

#check#
table(df10$mlb026)

#######################
#has children/freq. seeing (mlb009a)#
#######################
#check#
table(df10$mlb009a)

#code (<= 5) to (0); (6) to (1) per Shah et al. (2023)#
df10$mlb009a <- ifelse(df10$mlb009a < 6, 0, 1)

#check#
table(df10$mlb009a)

#######################
#activities w/ younger folks; grandchildren, etc. (mlb001b)#
#no change#
#######################
#check#
table(df10$mlb001b)

#######################
#work (mlb045)#
#######################
#check#
table(df10$mlb045)

#code (5) to (0), no change in (1) per Shah et al. (2023)#
df10$mlb045[df10$mlb045 == 5] <- 0
  
#check#
table(df10$mlb045)

#######################
#other volunteer/charity work (mlb001d)#
#######################
#check#
table(df10$mlb001d)

#code (<= 6) to (0); (7) to (1) per Shah et al. (2023)#
df10$mlb001d <- ifelse(df10$mlb001d < 7, 0, 1)

#check#
table(df10$mlb001d)

#######################
#feel isolated (mlb020c)#
#######################
#check#
table(df10$mlb020c)

#code (1) to (0); (2) to (1); (3) to (2) per Shah et al. (2023)#
df10$mlb020c <- df10$mlb020c - 1

#check#
table(df10$mlb020c)

#######################
#treated w/ less courtesy/respect (mlb030a)#
#######################
#check#
table(df10$mlb030a)

#code (<= 4) to (0); (5, 6) to (1) per Shah et al. (2023)# 
df10$mlb030a <- ifelse(df10$mlb030a < 5, 0, 1)

#check#
table(df10$mlb030a)
```


# Change variable names for modeling and visualization 

```{r}
sficolnames10 <- c("r10agey_e" = "age", 
                 "ragender" = "sex", 
                 "mlb021f" = "neighborhood", 
                 "mlb026" = "financial_control", 
                 "mlb009a" = "freq_children", 
                 "mlb001b" = "activities_youth", 
                 "mlb045" = "work", 
                 "mlb001d" = "volunteering", 
                 "mlb020c" = "feel_isolated", 
                 "mlb030a" = "courtresp")

colnames(df10)[colnames(df10) %in% names(sficolnames10)] <- 
  sficolnames10[colnames(df10)[colnames(df10) %in% names(sficolnames10)]]

colnames(df10oc)[colnames(df10oc) %in% names(sficolnames10)] <- 
  sficolnames10[colnames(df10oc)[colnames(df10oc) %in% names(sficolnames10)]]

#need to change labeling of sex for feature imp function#
if ("haven_labelled" %in% class(df10$sex)) {
  df10$sex <- as.numeric(df10$sex)
}
```


# Replicate GLM for mortality - SFI10 

```{r}
#######################
#specifying models#
#######################
sfi <- glm(mortality ~ age + sex + neighborhood + financial_control + freq_children 
           + activities_youth + work + volunteering + feel_isolated + courtresp, 
           family = binomial, data = df10, na.action = na.omit)

summary(sfi)

df10$SFIlp <- predict(sfi, type = "link")
df10$SFIpr <- predict(sfi, type = "response")

df10$SFIlpround <- round(df10$SFIlp, digits = 2)
df10$SFIprround <- round(df10$SFIpr, digits = 2)

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "2010SFIdistributions.pdf"), width = 6, height = 5)
hist(df10$SFIlp, breaks = "sturges", freq = FALSE, 
     xlim = c(-6, 4), ylim = c(0, 0.4), 
     main = "2010 Cohort: SFI-10", xlab = "Mortality Risk (Log Probability)")
lines(density(df10$SFIlp), col = "#27933E")
hist(df10$SFIpr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 5), 
     main = "2010 Cohort: SFI-10", xlab = "Mortality Risk (Probability)")
lines(density(df10$SFIpr), col = "#27933E")
dev.off()
```


# Age - Mortality GLM 

```{r}
#######################
#specifying models#
#######################
glmage <- glm(mortality ~ age, family = binomial, data = df10, na.action = na.omit)
summary(glmage)

df10$agelp <- predict(glmage, type = "link")
df10$agepr <- predict(glmage, type = "response")

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "2010AGEdistributions.pdf"), width = 6, height = 5)
hist(df10$agelp, breaks = "sturges", freq = FALSE, 
     xlim = c(-4, 2), ylim = c(0, 0.5), 
     main = "2010 Cohort: Age", xlab = "Mortality Risk (Log Probability)")
lines(density(df10$agelp), col = "#27933E")
hist(df10$agepr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 6), 
     main = "2010 Cohort: Age", xlab = "Mortality Risk (Probability)")
lines(density(df10$agepr), col = "#27933E")
dev.off()
```


# Age & Sex - Mortality GLM 

```{r}
#######################
#specifying models#
#######################
sexage <- glm(mortality ~ age + sex, family = binomial, data = df10, na.action = na.omit)
summary(sexage)

df10$sexagelp <- predict(sexage, type = "link")
df10$sexagepr <- predict(sexage, type = "response")

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "2010agesexdistributions.pdf"), width = 6, height = 5)
hist(df10$sexagelp, breaks = "sturges", freq = FALSE, 
     xlim = c(-4, 2), ylim = c(0, 0.6), 
     main = "2010 Cohort: Age & Sex", xlab = "Mortality Risk (Log Probability)")
lines(density(df10$sexagelp), col = "#27933E")
hist(df10$sexagepr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 5), 
     main = "2010 Cohort: Age & Sex", xlab = "Mortality Risk (Probability)")
lines(density(df10$sexagepr), col = "#27933E")
dev.off()
```


# Social items (SFI8) - Mortality GLM 

```{r}
#######################
#specifying models#
#######################
sfisoc <- glm(mortality ~ neighborhood + financial_control + freq_children 
           + activities_youth + work + volunteering + feel_isolated + courtresp, 
           family = binomial, data = df10, na.action = na.omit)

summary(sfisoc)

df10$socSFIlp <- predict(sfisoc, type = "link")
df10$socSFIpr <- predict(sfisoc, type = "response")

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "2010SFI8distributions.pdf"), width = 6, height = 5)
hist(df10$socSFIlp, breaks = "sturges", freq = FALSE, 
     xlim = c(-4, 2), ylim = c(0, 0.6), 
     main = "2010 Cohort: SFI-8", xlab = "Mortality Risk (Log Probability)")
lines(density(df10$socSFIlp), col = "#27933E")
hist(df10$socSFIpr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), 
     main = "2010 Cohort: SFI-8", xlab = "Mortality Risk (Probability)")
lines(density(df10$socSFIpr), col = "#27933E")
dev.off()
```


# Model evaluation - 2010 

### AIC/BIC 

```{r}
#all models specified#
hrsglmsBIC <- BIC(glmage, sexage, sfisoc, sfi)
hrsglmsAIC <- AIC(glmage, sexage, sfisoc, sfi)

hrsglmmodelfit <- cbind(hrsglmsBIC, hrsglmsAIC$AIC)

print(hrsglmmodelfit)
 
write.csv(hrsglmmodelfit, file.path(outdir, "2010ModelFit.csv"), row.names = TRUE)
```


### BCa bootstrap functions

```{r}
stratboot <- function(data, statistic, R = 9999, strata, conf.level = 0.95, ...) {
  levels <- unique(strata)
  n <- nrow(data)

  resamples <- replicate(R, {
    unlist(lapply(levels, function(lvl) {
      idx <- which(strata == lvl)
      sample(idx, length(idx), replace = TRUE)
    }))
  }, simplify = FALSE)

  statwrap <- function(data, indices) statistic(data[indices, ], ...)

  bootobj <- boot(data = data, statistic = statwrap, R = R, sim = "ordinary")
  bootobj$t <- sapply(resamples, function(i) statwrap(data, i))
  bootobj$t <- matrix(bootobj$t, ncol = 1)

  ci <- boot.ci(bootobj, type = "bca", conf = conf.level)

  list(
    estimate = statwrap(data, 1:n),
    bca_ci = ci$bca[4:5],
    bootobj = bootobj
  )
}

auroc <- function(data) {
  pROC::auc(pROC::roc(data$response, data$predicted, quiet = TRUE)) |> as.numeric()
}

auprc <- function(data) {
  PRROC::pr.curve(
    scores.class0 = data$predicted[data$response == 1],
    scores.class1 = data$predicted[data$response == 0],
    curve = FALSE)$auc.integral
}

aurocdiff <- function(data) {
  auc1 <- pROC::auc(pROC::roc(data$response, data$pred1, quiet = TRUE)) |> as.numeric()
  auc2 <- pROC::auc(pROC::roc(data$response, data$pred2, quiet = TRUE)) |> as.numeric()
  auc1 - auc2
}

auprcdiff <- function(data) {
  pr1 <- PRROC::pr.curve(scores.class0 = data$pred1[data$response == 1],
                         scores.class1 = data$pred1[data$response == 0],
                         curve = FALSE)$auc.integral
  pr2 <- PRROC::pr.curve(scores.class0 = data$pred2[data$response == 1],
                         scores.class1 = data$pred2[data$response == 0],
                         curve = FALSE)$auc.integral
  pr1 - pr2
}
```


### 2010 AUCs

```{r}
#######################
#SFI-10#
#######################
set.seed(555)
roccurvesfi10 <- roc(df10$mortality, df10$SFIpr, auc = TRUE, 
                     ci = TRUE, of = "auc", ci.method = "bootstrap", 
                     conf.level = 0.95, boot.n = 9999, boot.stratified = TRUE)
aucsfi10 <- roccurvesfi10$auc
ciaucsfi10 <-  roccurvesfi10$ci
print(aucsfi10)
print(ciaucsfi10)

dfbcarocsfi10 <- data.frame(response = df10$mortality, 
                            predicted = df10$SFIpr)
bcaciaucsfi10 <- stratboot(dfbcarocsfi10, auroc, R = 9999, strata = dfbcarocsfi10$response)
print(bcaciaucsfi10)

dfrocsfi10 <- data.frame(specificity = 1 - roccurvesfi10$specificities, 
                         sensitivity = roccurvesfi10$sensitivities)

pdf(file.path(outdir, "2010SFI10AUCROC.pdf"), width = 5, height = 5)
ggplot(dfrocsfi10, aes(x = specificity, y = sensitivity)) + 
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
    labs(title = paste0("2010 SFI-10: Mortality (AUC = ", round(aucsfi10, 3), ")"), 
         x = "1 - Specificity", y = "Sensitivity") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))
dev.off()


prcurvesfi10 <- pr.curve(scores.class0 = df10$SFIpr, weights.class0 = df10$mortality, curve = TRUE)
auprcsfi10 <- prcurvesfi10$auc.integral
print(auprcsfi10)

dfbcaprcsfi10 <- data.frame(response = df10$mortality, 
                            predicted = df10$SFIpr)
bcaciauprcsfi10 <- stratboot(dfbcarocsfi10, auprc, R = 9999, strata = dfbcaprcsfi10$response)
print(bcaciauprcsfi10)

dfprcsfi10 <- data.frame(recall = prcurvesfi10$curve[, 1], 
                         precision = prcurvesfi10$curve[, 2])

pdf(file.path(outdir, "2010SFI10AUCPRC.pdf"), width = 5, height = 5)
bline10 <- (sum(df10$mortality == TRUE)) / length(df10$mortality)
bline10
  
ggplot(dfprcsfi10, aes(x = recall, y = precision)) +
  geom_line(color = "#27933E", size = 1) +
   geom_abline(intercept = bline10, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = paste0("2010 SFI-10: Mortality (AUC = ", sprintf("%.3f", auprcsfi10), ")"),
    x = "Sensitivity", y = "PPV") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))
plot(prcurvesfi10)
dev.off()
```


# 2012 

# Completion of Form LB in 2012 & 65+ 

```{r}
#2012 completion#
c12 <- lrand[lrand$r11lbcomp == 1, ]
c12$lb_year <- 2012
c12 <- c12[!is.na(c12$r11lbcomp), ]

#N = 7412 (Can't find at https://github.com/sachinjshah/Social-Frailty-Index)#

#age 65+?#
#note: Shah et al. use "mage > 65", so technically 66 +#
#using r10agem_e for age (in months to use most restrictive unit)# 
#at END of interview (see RAND longitudinal p. 129).#
#Note: 66 * 12 = 792 so "> 791"#
c12 <- c12[c12$r11agem_e > 791, ]

#N = 3956 (3948 reported by Shah et al.)
```


# Create final df for 2012 cohort 

```{r}
#set wd#
setwd("/example/HRSfiles/")

#required for sample descriptive stats#
lranddescvars12 <- c("r11iwstat", "hacohort", 
                   "raracem", "rahispan", "raeduc", 
                   "r11shlt", "r11diabe", "r11cancre","r11hearte",
                   "r11stroke", "r11arthre", 
                   "r11smokev", "r11smoken",
                   "h11itot")

#required for sfi#
lrandsfivars12 <- c("hhidpn", "r11lbcomp", "lb_year", "ragender", "r11agey_e", 
               "r11agem_e", "r11iwbeg", "r11iwendm", "r11iwendy", 
               "radyear", "radmonth")

c12 <- c12[, c(lranddescvars12, lrandsfivars12)]

#convert r10iwbeg from SAS f8.0 date (days since 1/1/1960)#
#to year and month#
c12$r11iwbeg_r <- as.Date(c12$r11iwbeg, origin = "1960-01-01")

c12$iwbeg_y <- format(c12$r11iwbeg_r, "%Y")
c12$iwbeg_y <- as.numeric(c12$iwbeg_y)
c12$iwbeg_m <- format(c12$r11iwbeg_r, "%m")
c12$iwbeg_m <- as.numeric(c12$iwbeg_m)

#import Rand HRS 2012 fat file#
rand12 <- read_sav(paste0(getwd(), "/data/2012RAND/h12f3a_SPSS/h12f3a.sav"))
names(rand12) <- tolower(names(rand12))

clb12 <- merge(c12, rand12, by = "hhidpn", all.x = TRUE)

df12 <- clb12[, c(lranddescvars12, lrandsfivars12, "iwbeg_y", "iwbeg_m", "nd171", 
                  "nc095", "nc103", "nc104", "nc105",
                  "nlb021f", "nlb026", "nlb009a", "nlb001b", 
                  "nlb079", "nlb001d", "nlb020c", "nlb030a")]
```


# 4yr mortality

```{r}
#basing "4yr" mortality calculation on Shah et al.'s ACTUAL criteria (not 4yr mortality)#
#via authors: We mentioned "4-year mortality" in the paper, but the way we defined the#
#death_flag (the outcome, a binary indicator) was to include all deaths that occurred by#
#the end of 2017 (Dec 31, 2017) for the 2012 cohort.#

#note: dates cannot be narrowed down from months#
#note: <= 2018 because December, 2017 would be 2017 + 1#
#in the formula below, so 2018#

df12$DOD <- df12$radyear + df12$radmonth / 12
df12$mortality <- ifelse(is.na(df12$DOD), 0,
                         ifelse(df12$DOD <= 2018, 1, 0))

table(df12$mortality == 1)
#906/3956 compared to 882/3948 published#

#just checking when died#
df12$interview <- df12$r11iwendy + df12$r11iwendm / 12
df12$mortalitytime <- df12$DOD - df12$interview

summary(df12$mortalitytime[df12$mortality == 1])
hist(df12$mortalitytime[df12$mortality == 1], breaks = "sturges", freq = FALSE, 
     main = "2012 Cohort: Mortality Time", xlab = "Time (Years)")
lines(density(df12$mortalitytime[df12$mortality == 1]), col = "#27933E")

#just checking outcome assessment window#
df12$outcometime <- 2018 - df12$interview
summary(df12$outcometime)
```

Total 4yr mortality suggested = 906
Max mortality time = 5.667 years
Shah et al. (2023) PNAS reported = 882


# 2012 Descriptive Stats

The following 2 sections can be combined to recreate the descriptive statistics in Table S1.

### Descriptive stats - demographics, health status, function

```{r}
#####################
#demographics#
#####################
#age (median+IQR)#
summary(df12$r11agey_e)
hist(df12$r11agey_e, breaks = "sturges", freq = FALSE,
     main = "2012 Cohort: Age", xlab = "Age (Years)")
lines(density(df12$r11agey_e), col = "#27933E")

#sex (female+%)#
table(df12$ragender == 2)


#####################
#health status#
#####################
#fair or poor self reported health (N+%)#
table(!is.na(df12$r11shlt))
table(df12$r11shlt  %in% c(4, 5))

#diabetes (N+%)#
table(df12$r11diabe == 1)

#cancer (excluding minor skin cancer) (N+%)#
table(df12$r11cancre == 1)

#heart disease (N+%)#
table(df12$r11hearte == 1)

#stroke (N+%)#
table(df12$r11stroke == 1)

#arthritis (N+%)#
table(df12$r11arthre == 1)

#current smoker? (N+%)#
table(df12$r11smoken == 1)

#former smoker? (N+%)#
table(df12$r11smokev == 1 & df12$r11smoken == 0)


#####################
#mortality#
#####################
#mortality#
table(df12$mortality == 1)


#####################
#summarize#
#####################
N <- 3956

# Generate table with relevant variables
c12stats <- data.frame(
  Variable = c(
    "Fair or Poor Self-Reported Health",
    "Diabetes",
    "Cancer (excluding minor skin cancer)",
    "Heart Disease",
    "Stroke",
    "Arthritis",
    "Current Smoker",
    "Former Smoker"
  ),
  `N (%)` = c(
    freq_percent(df12$r11shlt, c(4, 5)),
    freq_percent(df12$r11diabe, 1),  
    freq_percent(df12$r11cancre, 1),  
    freq_percent(df12$r11hearte, 1),  
    freq_percent(df12$r11stroke, 1),  
    freq_percent(df12$r11arthre, 1),  
    freq_percent(df12$r11smoken, 1),  
    freq_percent(df12$r11smokev[df12$r11smoken == 0], 1)
  )
)

write.csv(c12stats, file.path(outdir, "2012 HRS health and function.csv"), row.names = TRUE)
```


### Descriptive stats - social characteristics 

```{r}
#####################
#general#
#####################
#race (N+%)#
table(df12$raracem)
table(!is.na(df12$raracem))

#ethnicity (N+%)#
table(df12$rahispan)
table(!is.na(df12$rahispan))

#education (N+%)#
table(df12$raeduc)
table(!is.na(df12$raeduc))

#missing 1; Shah et al. (2023) are too#

#income (N+%)#
summary(df12$h11itot)
hist(df12$h11itot, breaks = "sturges", freq = FALSE, 
     xaxt = 'n', ylim = c(0, 0.000020),
     main = "2010 Cohort: Household Income", xlab = "Annual Income (USD)")
axis(side = 1, at = pretty(range(df12$h11itot)), labels = pretty(range(df12$h11itot)))
lines(density(df12$h11itot), col = "#27933E")

#####################
#sfi vars#
#####################
#all tables based on functional forms analysis below#
#neighborhood unclean#
#code (< 4) to (0); (>= 4) to (1)#
table(df12$nlb021f >= 4)

#low financial control#
#code (< 2) to (0); (>= 2) to (1)#
table(df12$nlb026 < 2)

#meets with children less than once/year#
#code (<= 5) to (0); (6) to (1)#
table(df12$nlb009a == 6)

#activities w/ youth#
summary(df12$nlb001b)

#work#
table(df12$nlb079 == 1)

#no volunteer#
#code (<= 6) to (0); (7) to (1)#
table(df12$nlb001d == 7)

#loneliness#
table(df12$nlb020c)

#courtesy/respect#
#code (<= 4) to (0); (5, 6) to (1)#
table(df12$nlb030a <= 4)

#####################
#summarize#
#####################
N = 3956

c12descstats <- data.frame(
  Variable = c(
    "Black",
    "White",
    "Other", 
    "Hispanic origin", 
    "Not of hispanic origin", 
    "Less than high school", 
    "High school diploma or equiv", 
    "Some college", 
    "College and above",
    "< $56,599", 
    "$56,600 - 169,799", 
    "> $169,800", 
    "Age", 
    "Sex", 
    "Neighborhood is not clean", 
    "Low percieved financial control", 
    "Meets with children less than once per year or equivalent", 
    "Activities with younger folks or equivalent", 
    "Presently working", 
    "Does no volunteer/charity work", 
    "Feels isolated often", 
    "Feels isolated some of the time", 
    "Feels isolated hardly ever or never",
    "Often treated with less courtesy and respect than others"
  ),
  `N (%)` = c(
    freq_percent(df12$raracem, 2),
    freq_percent(df12$raracem, 1),
    freq_percent(df12$raracem, 3),
    freq_percent(df12$rahispan, 1), 
    freq_percent(df12$rahispan, 0),
    freq_percent(df12$raeduc, 1),
    freq_percent(df12$raeduc, 2:3),
    freq_percent(df12$raeduc, 4),
    freq_percent(df12$raeduc, 5),
    freq_percentlog(df12$h11itot, df12$h11itot < 56600),
    freq_percentlog(df12$h11itot, df12$h11itot >= 56600 & df12$h11itot <= 169799),
    freq_percentlog(df12$h11itot, df12$h11itot > 169799),
    paste0(round(median(df12$r11agey_e), 2), 
           " [", round(summary(df12$r11agey_e)[2], 2),",", 
           round(summary(df12$r11agey_e)[5], 2), "]"), 
    freq_percent(df12$ragender, 2), 
    freq_percent(df12$nlb021f, 4:10), 
    freq_percent(df12$nlb026, 0:1), 
    freq_percent(df12$nlb009a, 6), 
    paste0(round(median(df12$nlb001b), 2), 
           " [", round(summary(df12$nlb001b)[2], 2), ",",
           round(summary(df12$nlb001b)[5], 2), "]"),
    freq_percent(df12$nlb079, 1), 
    freq_percent(df12$nlb001d, 7), 
    freq_percent(df12$nlb020c, 1), 
    freq_percent(df12$nlb020c, 2),
    freq_percent(df12$nlb020c, 3),
    freq_percent(df12$nlb030a, 1:4)
  )
)

write.csv(c12descstats, file.path(outdir, "2012 HRS descriptive stats.csv"), row.names = TRUE)
```


# Single imputation for missing values 

### find mode & median

```{r}
#######################
#age (r11agey_e)#
#######################
median(df12$r11agey_e, na.rm = TRUE)

#median: 75#

#######################
#gender (ragender)#
#######################
findmode(df12$ragender)

#mode: 2 (woman)#

#######################
#neighborhood clean (nlb021f)#
#######################
findmode(df12$nlb021f)

#mode: 1 (highest agreement with "this area is kept very clean")#

#######################
#financial control (nlb026)#
#######################
findmode(df12$nlb026)

#mode: 10 (very much control)#

#######################
#has children/freq. seeing (nlb009a)#
#######################
findmode(df12$nlb009a)

#mode: 2 (once or twice a week)#

#######################
#activities w/ younger folks; grandchildren, etc. (nlb001b)#
#######################
findmode(df12$nlb001b)

#mode: 7 (never/not relevant)#

#######################
#work (nlb079)#
#######################
findmode(df12$nlb079)

#mode: 5 (No)#

#######################
#other volunteer/charity work (nlb001d)#
#######################
findmode(df12$nlb001d)

#mode: 7 (never/not relevant)#

#######################
#feel isolated (nlb020c)#
#######################
findmode(df12$nlb020c)

#mode: 3 (hardly ever or never)#

#######################
#treated w/ less courtesy/respect (nlb030a)#
#######################
findmode(df12$nlb030a)

#mode: 6 (never)#
```


### Single imputation of median and mode for missing values

```{r}
#######################
#age (r11agey_e)#
#######################
table(!is.na(df12$r11agey_e))
df12$r11agey_e[is.na(df12$r11agey_e)] <- 75
table(!is.na(df12$r11agey_e))

#######################
#sex (ragender)#
#######################
table(!is.na(df12$ragender))
df12$ragender[is.na(df12$ragender)] <- 2
table(!is.na(df12$ragender))

#######################
#neighborhood clean (nlb021f)#
#######################
table(!is.na(df12$nlb021f))
df12$nlb021f[is.na(df12$nlb021f)] <- 1
table(!is.na(df12$nlb021f))

#######################
#financial control (nlb026)#
#######################
table(!is.na(df12$nlb026))
df12$nlb026[is.na(df12$nlb026)] <- 10
table(!is.na(df12$nlb026))

#######################
#has children/freq. seeing (nlb009a)#
#######################
table(!is.na(df12$nlb009a))
df12$nlb009a[is.na(df12$nlb009a)] <- 2
table(!is.na(df12$nlb009a))

#######################
#activities w/ younger folks; grandchildren, etc. (nlb001b)#
#######################
table(!is.na(df12$nlb001b))
df12$nlb001b[is.na(df12$nlb001b)] <- 7
table(!is.na(df12$nlb001b))

#######################
#work (nlb079)#
#######################
table(!is.na(df12$nlb079))
df12$nlb079[is.na(df12$nlb079)] <- 5
table(!is.na(df12$nlb079))

#######################
#other volunteer/charity work (nlb001d)#
#######################
table(!is.na(df12$nlb001d))
df12$nlb001d[is.na(df12$nlb001d)] <- 7
table(!is.na(df12$nlb001d))

#######################
#feel isolated (nlb020c)#
#######################
table(!is.na(df12$nlb020c))
df12$nlb020c[is.na(df12$nlb020c)] <- 3
table(!is.na(df12$nlb020c))

#######################
#treated w/ less courtesy/respect (nlb030a)#
#######################
table(!is.na(df12$nlb030a))
df12$nlb030a[is.na(df12$nlb030a)] <- 6
table(!is.na(df12$nlb030a))
```


# Recode vars based on functional forms 

```{r}
#######################
#create dummy df for cors later#
#######################
df12oc <- df12

#######################
#age (r11agey_e)#
#no change in Shah et al. (2023)#
#######################

#######################
#gender (ragender)#
#######################
#check#
table(df12$ragender)

#code male (1) to (0); female (2) to (1) per Shah et al. (2023)#
df12$ragender[df12$ragender == 1] <- 0
df12$ragender[df12$ragender == 2] <- 1

#check#
table(df12$ragender)

#######################
#neighborhood clean? (nlb021f)#
#######################
#check#
table(df12$nlb021f)

#code (< 4) to (0); (>= 4) to (1) per Shah et al. (2023)#
df12$nlb021f <- ifelse(df12$nlb021f < 4, 0, 1)

#check#
table(df12$nlb021f)

#######################
#financial control (nlb026)#
#######################
#check#
table(df12$nlb026)

#code (< 2) to (0); (>= 2) to (1) per my manual categorization#
df12$nlb026 <- ifelse(df12$nlb026 < 2, 0, 1)

#check#
table(df12$nlb026)

#######################
#has children/freq. seeing (nlb009a)#
#######################
#check#
table(df12$nlb009a)

#code (<= 5) to (0); (6) to (1) per Shah et al. (2023)#
df12$nlb009a <- ifelse(df12$nlb009a < 6, 0, 1)

#check#
table(df12$nlb009a)

#######################
#activities w/ younger folks; grandchildren, etc. (nlb001b)#
#no change#
#######################
#check#
table(df12$nlb001b)

#######################
#work (nlb079)#
#######################
#check#
table(df12$nlb079)

#code (5) to (0), no change in (1) per Shah et al. (2023)#
df12$nlb079[df12$nlb079 == 5] <- 0
  
#check#
table(df12$nlb079)

#######################
#other volunteer/charity work (nlb001d)#
#######################
#check#
table(df12$nlb001d)

#code (<= 6) to (0); (7) to (1) per Shah et al. (2023)#
df12$nlb001d <- ifelse(df12$nlb001d < 7, 0, 1)

#check#
table(df12$nlb001d)

#######################
#feel isolated (nlb020c)#
#######################
#check#
table(df12$nlb020c)

#code (1) to (0); (2) to (1); (3) to (2) per Shah et al. (2023)#
df12$nlb020c <- df12$nlb020c - 1

#check#
table(df12$nlb020c)

#######################
#treated w/ less courtesy/respect (nlb030a)#
#######################
#check#
table(df12$nlb030a)

#code (<= 4) to (0); (5, 6) to (1) per Shah et al. (2023)# 
df12$nlb030a <- ifelse(df12$nlb030a < 5, 0, 1)

#check#
table(df12$nlb030a)
```


# Change var names for modeling and viz 

```{r}
sficolnames12 <- c("r11agey_e" = "age", 
                 "ragender" = "sex", 
                 "nlb021f" = "neighborhood", 
                 "nlb026" = "financial_control", 
                 "nlb009a" = "freq_children", 
                 "nlb001b" = "activities_youth", 
                 "nlb079" = "work", 
                 "nlb001d" = "volunteering", 
                 "nlb020c" = "feel_isolated", 
                 "nlb030a" = "courtresp")

colnames(df12)[colnames(df12) %in% names(sficolnames12)] <- 
  sficolnames12[colnames(df12)[colnames(df12) %in% names(sficolnames12)]]

colnames(df12oc)[colnames(df12oc) %in% names(sficolnames12)] <- 
  sficolnames12[colnames(df12oc)[colnames(df12oc) %in% names(sficolnames12)]]

#need to change labeling of sex for feature imp funciton#
if ("haven_labelled" %in% class(df12$sex)) {
  df12$sex <- as.numeric(df12$sex)
}
```


# Apply SFI10 coeffs from 2010

```{r}
#######################
#specifying models#
#######################
df12$SFIlp <- predict(sfi, newdata = df12, type = "link")
df12$SFIpr <- predict(sfi, newdata = df12, type = "response")

df12$SFIlpround <- round(df12$SFIlp, digits = 2)
df12$SFIprround <- round(df12$SFIpr, digits = 2)


#######################
#distribution viz#
#######################
pdf(file.path(outdir, "2012SFI10distributions.pdf"), width = 6, height = 5)
hist(df12$SFIlp, breaks = "sturges", freq = FALSE, 
     xlim = c(-6, 4), ylim = c(0, 0.4), 
     main = "2012 Cohort: SFI-10", xlab = "Mortality Risk (Log Probability)")
lines(density(df12$SFIlp), col = "#27933E")
hist(df12$SFIpr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 6), 
     main = "2012 Cohort: SFI-10", xlab = "Mortality Risk (Probability)")
lines(density(df12$SFIpr), col = "#27933E")
dev.off()
```


# Apply ageglm coeffs from 2010 

```{r}
#######################
#specifying models#
#######################
df12$agelp <- predict(glmage, newdata = df12, type = "link")
df12$agepr <- predict(glmage, newdata = df12, type = "response")

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "2012AGEdistributions.pdf"), width = 6, height = 5)
hist(df12$agelp, breaks = "sturges", freq = FALSE, 
     xlim = c(-4, 2), ylim = c(0, 0.6), 
     main = "2010 Cohort: Age", xlab = "Mortality Risk (Log Probability)")
lines(density(df12$agelp), col = "#27933E")
hist(df12$agepr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 6), 
     main = "2010 Cohort: Age", xlab = "Mortality Risk (Probability)")
lines(density(df12$agepr), col = "#27933E")
dev.off()
```


# Apply age & sex coeffs from 2010 

```{r}
#######################
#specifying models#
#######################
df12$sexagelp <- predict(sexage, newdata = df12, type = "link")
df12$sexagepr <- predict(sexage, newdata = df12, type = "response")

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "2012agesexdistributions.pdf"), width = 6, height = 5)
hist(df12$sexagelp, breaks = "sturges", freq = FALSE, 
     xlim = c(-4, 2), ylim = c(0, 0.6), 
     main = "2012 Cohort: Age & Sex", xlab = "Mortality Risk (Log Probability)")
lines(density(df12$sexagelp), col = "#27933E")
hist(df12$sexagepr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 6), 
     main = "2012 Cohort: Age & Sex", xlab = "Mortality Risk (Probability)")
lines(density(df12$sexagepr), col = "#27933E")
dev.off()
```


# Apply SFI8 coeffs from 2010

```{r}
#######################
#specifying models#
#######################
df12$socSFIlp <- predict(sfisoc, newdata = df12, type = "link")
df12$socSFIpr <- predict(sfisoc, newdata = df12, type = "response")

#######################
#distribution viz#
#######################
pdf(file.path(outdir, "2012SFI8distributions.pdf"), width = 6, height = 5)
hist(df12$socSFIlp, breaks = "sturges", freq = FALSE, 
     xlim = c(-5, 2), ylim = c(0, 0.6), 
     main = "2012 Cohort: SFI-8", xlab = "Mortality Risk (Log Probability)")
lines(density(df12$socSFIlp), col = "#27933E")
hist(df12$socSFIpr, breaks = "sturges", freq = FALSE, 
     xlim = c(0, 1), ylim = c(0, 6), 
     main = "2012 Cohort: SFI-8", xlab = "Mortality Risk (Probability)")
lines(density(df12$socSFIpr), col = "#27933E")
dev.off()
```


# Model evaluation - 2012 

### 2012 AUC ROC mortality 

```{r}
#######################
#SFI - my replication (sfi)#
#######################
set.seed(555)
roccurvesfi12 <- roc(df12$mortality, df12$SFIpr, auc = TRUE, 
                     ci = TRUE, of = "auc", ci.method = "bootstrap", 
                     conf.level = 0.95, boot.n = 9999, boot.stratified = TRUE)
aucsfi12 <- roccurvesfi12$auc
ciaucsfi12 <-  roccurvesfi12$ci
print(aucsfi12)
print(ciaucsfi12)

dfbcarocsfi12 <- data.frame(response = df12$mortality, 
                            predicted = df12$SFIpr)
bcaciaucsfi12 <- stratboot(dfbcarocsfi12, auroc, R = 9999, strata = dfbcarocsfi12$response)
print(bcaciaucsfi12)

dfrocsfi12 <- data.frame(specificity = 1 - roccurvesfi12$specificities, 
                    sensitivity = roccurvesfi12$sensitivities)

pdf(file.path(outdir, "2012AUCROCs.pdf"), width = 4, height = 4)
ggplot(dfrocsfi12, aes(x = specificity, y = sensitivity)) + 
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
    labs(title = paste0("Mortality (AUC = ", sprintf("%.3f", aucsfi12), ")"), 
         x = "1 - Specificity", y = "Sensitivity") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))


#######################
#age#
#######################
roccurveage12 <- roc(df12$mortality, df12$agepr, auc = TRUE, 
                     ci = TRUE, of = "auc", ci.method = "bootstrap", 
                     conf.level = 0.95, boot.n = 9999, boot.stratified = TRUE)
aucage12 <- roccurveage12$auc
ciaucage12 <-  roccurveage12$ci
print(aucage12)
print(ciaucage12)

dfbcarocage12 <- data.frame(response = df12$mortality, 
                            predicted = df12$agepr)
bcaciaucage12 <- stratboot(dfbcarocage12, auroc, R = 9999, strata = dfbcarocage12$response)
print(bcaciaucage12)

dfrocage12 <- data.frame(specificity = 1 - roccurveage12$specificities, 
                    sensitivity = roccurveage12$sensitivities)

ggplot(dfrocage12, aes(x = specificity, y = sensitivity)) + 
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
    labs(title = paste0("Age: Mortality (AUC = ", sprintf("%.3f",aucage12), ")"), 
         x = "1 - Specificity", y = "Sensitivity") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))


#######################
#age & sex#
#######################
roccurvesexage12 <- roc(df12$mortality, df12$sexagepr, auc = TRUE, 
                     ci = TRUE, of = "auc", ci.method = "bootstrap", 
                     conf.level = 0.95, boot.n = 9999, boot.stratified = TRUE)
aucsexage12 <- roccurvesexage12$auc
ciaucsexage12 <-  roccurvesexage12$ci
print(aucsexage12)
print(ciaucsexage12)

dfbcarocsexage12 <- data.frame(response = df12$mortality, 
                               predicted = df12$sexagepr)
bcaciaucsexage12 <- stratboot(dfbcarocsexage12, auroc, R = 9999, strata = dfbcarocsexage12$response)
print(bcaciaucsexage12)

dfrocsexage12 <- data.frame(specificity = 1 - roccurvesexage12$specificities, 
                    sensitivity = roccurvesexage12$sensitivities)

ggplot(dfrocsexage12, aes(x = specificity, y = sensitivity)) + 
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
    labs(title = paste0("Age & Sex: Mortality (AUC = ", sprintf("%.3f",aucsexage12), ")"), 
         x = "1 - Specificity", y = "Sensitivity") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))


#######################
#social items#
#######################
roccurvesfisoc12 <- roc(df12$mortality, df12$socSFIpr, auc = TRUE, 
                     ci = TRUE, of = "auc", ci.method = "bootstrap", 
                     conf.level = 0.95, boot.n = 9999, boot.stratified = TRUE)
aucsfisoc12 <- roccurvesfisoc12$auc
ciaucsfisoc12 <-  roccurvesfisoc12$ci
print(aucsfisoc12)
print(ciaucsfisoc12)

dfbcarocsfisoc12 <- data.frame(response = df12$mortality, 
                               predicted = df12$socSFIpr)
bcaciaucsfisoc12 <- stratboot(dfbcarocsfisoc12, auroc, R = 9999, strata = dfbcarocsfisoc12$response)
print(bcaciaucsfisoc12)

dfrocsfisoc12 <- data.frame(specificity = 1 - roccurvesfisoc12$specificities, 
                    sensitivity = roccurvesfisoc12$sensitivities)

ggplot(dfrocsfisoc12, aes(x = specificity, y = sensitivity)) + 
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
    labs(title = paste0("SFI-8: Mortality (AUC = ", sprintf("%.3f",aucsfisoc12), ")"), 
         x = "1 - Specificity", y = "Sensitivity") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))


#######################
#combined plot#
#######################
dfrocsfi12$model <- paste0("SFI-10 (", sprintf("%.3f", aucsfi12), ")")
dfrocage12$model <- paste0("Age (", sprintf("%.3f", aucage12), ")")
dfrocsexage12$model <- paste0("Age & Sex (", sprintf("%.3f", aucsexage12), ")")
dfrocsfisoc12$model <- paste0("SFI-8 (", sprintf("%.3f", aucsfisoc12), ")")

dfrocs12 <- rbind(dfrocsfi12, dfrocage12, dfrocsexage12, dfrocsfisoc12)

dfrocs12$model <- factor(dfrocs12$model, levels = c(
  paste0("SFI-10 (", sprintf("%.3f", aucsfi12), ")"),
  paste0("Age & Sex (", sprintf("%.3f", aucsexage12), ")"),
  paste0("Age (", sprintf("%.3f", aucage12), ")"),
  paste0("SFI-8 (", sprintf("%.3f", aucsfisoc12), ")")))

modcols <- setNames(c("#27933E", "#B8CDF8", "#6A4C93", "#E4CC37"),
                    c(paste0("SFI-10 (", sprintf("%.3f", aucsfi12), ")"),
                      paste0("Age (", sprintf("%.3f", aucage12), ")"),
                      paste0("Age & Sex (", sprintf("%.3f", aucsexage12), ")"),
                      paste0("SFI-8 (", sprintf("%.3f", aucsfisoc12), ")")))


ggplot(dfrocs12, aes(x = specificity, y = sensitivity, color = model)) +
  geom_line(size = 1) +
  scale_color_manual(values = modcols) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
  labs(title = "Figure 1A with legend",
       x = "1 - Specificity", y = "Sensitivity", color = expression("Model (AU"[ROC]*")")) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 11),
    axis.text = element_text(size = 11), legend.title = element_text(size = 11),
    legend.text = element_text(size = 11), 
    legend.position = "right")

ggplot(dfrocs12, aes(x = specificity, y = sensitivity, color = model)) +
  geom_line(size = 1) +
  scale_color_manual(values = modcols) +
  geom_abline(intercept = 0, slope = 1, linetype = 5, color = "#5F6062") +
  labs(title = "Figure 1A",
       x = "1 - Specificity", y = "Sensitivity", color = expression("Model (AU"[ROC]*")")) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 11),
    axis.text = element_text(size = 11), legend.title = element_text(size = 11),
    legend.text = element_text(size = 11), 
    legend.position = "none")

#######################
#testing#
#######################
#BCa CI-based bootstrap test: SexAge vs. SFI-10#
dfdiffsexage <- data.frame(response = df12$mortality,
                           pred1 = df12$SFIpr,
                           pred2 = df12$sexagepr)

sfisexagediff <- stratboot(dfdiffsexage, aurocdiff, R = 9999, strata = dfdiffsexage$response, conf.level = 0.90)
print(sfisexagediff)
```


### 2012 AUC PRC mortality

```{r}
#######################
#baseline#
#######################
bline12 <- (sum(df12$mortality == TRUE)) / length(df12$mortality)
bline12

#######################
#SFI#
#######################
set.seed(555)
prcurvesfi12 <- pr.curve(scores.class0 = df12$SFIpr, weights.class0 = df12$mortality, curve = TRUE)
auprcsfi12 <- prcurvesfi12$auc.integral
print(auprcsfi12)

dfbcaprcsfi12 <- data.frame(response = df12$mortality, 
                            predicted = df12$SFIpr)
bcaciauprcsfi12 <- stratboot(dfbcarocsfi12, auprc, R = 9999, strata = dfbcaprcsfi12$response)
print(bcaciauprcsfi12)

dfprcsfi12 <- data.frame(recall = prcurvesfi12$curve[, 1], 
                         precision = prcurvesfi12$curve[, 2])

pdf(file.path(outdir, "2012AUCPRCs.pdf"), width = 4, height = 4)
ggplot(dfprcsfi12, aes(x = recall, y = precision)) +
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = bline12, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = paste0("SFI-10: 4y Mortality (AUC = ", sprintf("%.3f", auprcsfi12), ")"),
    x = "Sensitivity", y = "PPV") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))

plot(prcurvesfi12)

#######################
#age#
#######################
prcurveage12 <- pr.curve(scores.class0 = df12$agepr, weights.class0 = df12$mortality, curve = TRUE)
auprcage12 <- prcurveage12$auc.integral
print(auprcage12)

dfbcaprcage12 <- data.frame(response = df12$mortality, 
                            predicted = df12$agepr)
bcaciauprcage12 <- stratboot(dfbcarocage12, auprc, R = 9999, strata = dfbcaprcage12$response)
print(bcaciauprcage12)

dfprcage12 <- data.frame(recall = prcurveage12$curve[, 1], 
                         precision = prcurveage12$curve[, 2])

ggplot(dfprcage12, aes(x = recall, y = precision)) +
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = bline12, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = paste0("Age: 4y Mortality (AUC = ", sprintf("%.3f", auprcage12), ")"),
    x = "Sensitivity", y = "PPV") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))

plot(prcurveage12)

#######################
#age & sex#
#######################
prcurvesexage12 <- pr.curve(scores.class0 = df12$sexagepr, weights.class0 = df12$mortality, curve = TRUE)
auprcsexage12 <- prcurvesexage12$auc.integral
print(auprcsexage12)

dfbcaprcsexage12 <- data.frame(response = df12$mortality, 
                               predicted = df12$sexagepr)
bcaciauprcsexage12 <- stratboot(dfbcarocsexage12, auprc, R = 9999, strata = dfbcaprcsexage12$response)
print(bcaciauprcsexage12)

dfprcsexage12 <- data.frame(recall = prcurvesexage12$curve[, 1], 
                         precision = prcurvesexage12$curve[, 2])

ggplot(dfprcsexage12, aes(x = recall, y = precision)) +
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = bline12, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = paste0("Age & Sex: 4y Mortality (AUC = ", sprintf("%.3f", auprcsexage12), ")"),
    x = "Sensitivity", y = "PPV") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))

plot(prcurvesexage12)

#######################
#social items#
#######################
prcurvesocsfi12 <- pr.curve(scores.class0 = df12$socSFIpr, weights.class0 = df12$mortality, curve = TRUE)
auprcsocsfi12 <- prcurvesocsfi12$auc.integral
print(auprcsocsfi12)

dfbcaprcsocsfi12 <- data.frame(response = df12$mortality, 
                               predicted = df12$socSFIpr)
bcaciauprcsocsfi12 <- stratboot(dfbcaprcsocsfi12, auprc, R = 9999, strata = dfbcaprcsocsfi12$response)
print(bcaciauprcsocsfi12)

dfprcsocsfi12 <- data.frame(recall = prcurvesocsfi12$curve[, 1], 
                         precision = prcurvesocsfi12$curve[, 2])

ggplot(dfprcsocsfi12, aes(x = recall, y = precision)) +
  geom_line(color = "#27933E", size = 1) +
  geom_abline(intercept = bline12, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = paste0("SFI-8: 4y Mortality (AUC = ", sprintf("%.3f", auprcsocsfi12), ")"),
    x = "Sensitivity", y = "PPV") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))

plot(prcurvesocsfi12)

#######################
#combined plot#
#######################
dfprcsfi12$model <- paste0("SFI-10 (", sprintf("%.3f", auprcsfi12), ")")
dfprcage12$model <- paste0("Age (", sprintf("%.3f", auprcage12), ")")
dfprcsexage12$model <- paste0("Age & Sex (", sprintf("%.3f", auprcsexage12), ")")
dfprcsocsfi12$model <- paste0("SFI-8 (", sprintf("%.3f", auprcsocsfi12), ")")

dfprcs12 <- rbind(dfprcsfi12, dfprcage12, dfprcsexage12, dfprcsocsfi12)

dfprcs12$model <- factor(dfprcs12$model, levels = c(
  paste0("SFI-10 (", sprintf("%.3f", auprcsfi12), ")"),
  paste0("Age & Sex (", sprintf("%.3f", auprcsexage12), ")"),
  paste0("Age (", sprintf("%.3f", auprcage12), ")"),
  paste0("SFI-8 (", sprintf("%.3f", auprcsocsfi12), ")")))

modcols <- setNames(c("#27933E", "#B8CDF8", "#6A4C93", "#E4CC37"),
                    c(paste0("SFI-10 (", sprintf("%.3f", auprcsfi12), ")"),
                      paste0("Age (", sprintf("%.3f", auprcage12), ")"),
                      paste0("Age & Sex (", sprintf("%.3f", auprcsexage12), ")"),
                      paste0("SFI-8 (", sprintf("%.3f", auprcsocsfi12), ")")))


ggplot(dfprcs12, aes(x = recall, y = precision, color = model)) +
  geom_line(size = 1) +
  scale_color_manual(values = modcols) +
  geom_abline(intercept = bline12, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = "Supplemental Figure 1A with legend",
       x = "Sensitivity", y = "PPV", color = expression("Model (AU"[PRC]*")")) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),
    axis.text = element_text(size = 13), legend.title = element_text(size = 13),
    legend.text = element_text(size = 12), 
    legend.position = "right")

ggplot(dfprcs12, aes(x = recall, y = precision, color = model)) +
  geom_line(size = 1) +
  scale_color_manual(values = modcols) +
  geom_abline(intercept = bline12, slope = 0, linetype = 5, color = "#5F6062") +
  labs(title = "Supplemental Figure 1A",
       x = "Sensitivity", y = "PPV", color = expression("Model (AU"[PRC]*")")) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),
    axis.text = element_text(size = 13), legend.title = element_text(size = 13),
    legend.text = element_text(size = 12), 
    legend.position = "none")

dev.off()

#######################
#testing#
#######################
#BCa CI-based bootstrap test: SexAge vs. SFI replication#
sfisexageprcdiff <- stratboot(dfdiffsexage, auprcdiff, R = 9999, strata = dfdiffsexage$response, conf.level = 0.90)
print(sfisexageprcdiff)
```


# 2012 Feature importance


```{r}
sfinames <- c("age", "sex", "neighborhood", "financial_control", 
              "freq_children", "activities_youth", "work", "volunteering", 
              "feel_isolated", "courtresp")

sfisocnames <- c("neighborhood", "financial_control", 
                 "freq_children", "activities_youth", "work", "volunteering", 
                 "feel_isolated", "courtresp")

aurocloss <- function(response, predicted) {
  rocobj <- roc(response, predicted)
  return(1 - rocobj$auc)
}

auprcloss <- function(response, predicted) {
  pc <- predicted[response == 1]
  nc <- predicted[response == 0]
  prcobj <- pr.curve(scores.class0 = pc, scores.class1 = nc)
  return(1 - prcobj$auc.integral)
}
```


### ROC-wise

```{r}
#######################
#SFI-10#
#######################
set.seed(555)
sfipred12 <- Predictor$new(model = sfi, data = df12[, sfinames], y = df12$mortality)

sfiimp12 <- FeatureImp$new(sfipred12, loss = aurocloss, compare = "difference", n.repetitions = 1000)

sfiimpplot12 <- plot(sfiimp12)

df12labs <- c(age = "Age", sex = "Sex", neighborhood = "Neighborhood clean", 
              financial_control = "Financial control", freq_children = "Meets w/ children",
              activities_youth = "Activities w/ youth", work = "Working", volunteering = "Volunteering", 
              feel_isolated = "Feeling isolated", courtresp = "Courtesy/respect")

pdf(file.path(outdir, "2012FeatureImpsROC.pdf"), width = 5.1, height = 4)
sfiimpplot12 +
  ggtitle("Figure 1B") +
  scale_x_continuous(name = expression("Feature Importance (Loss: 1 - AU"[ROC]*")"), 
                     limits = c(-0.01, 0.18)) +
  scale_y_discrete(name = NULL, labels = df12labs) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13.5),      
    axis.text = element_text(size = 13.5))

sfiimp12res <- sfiimp12$results
sfiimp12res
write.csv(sfiimp12res, file.path(outdir, "HRS 2012 SFI10 FeatureImpROC.csv"), row.names = TRUE)

#######################
#SFI-8#
#######################
socpred12 <- Predictor$new(model = sfisoc, data = df12[, sfisocnames], y = df12$mortality)

socimp12 <- FeatureImp$new(socpred12, loss = aurocloss, compare = "difference", n.repetitions = 1000)

socimpplot12 <- plot(socimp12)

socimpplot12 +
  ggtitle("Supplemental Figure 1B") +
  scale_x_continuous(name = expression("Feature Importance (Loss: 1 - AU"[ROC]*")")) +
  scale_y_discrete(name = NULL, labels = df12labs) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13.5),      
    axis.text = element_text(size = 13.5))

dev.off()

socimp12res <- socimp12$results
socimp12res
write.csv(socimp12res, file.path(outdir, "HRS 2012 SFI8 FeatureImpROC.csv"), row.names = TRUE)
```


### PRC-wise ###

```{r}
#######################
#SFI-10#
#######################
set.seed(555)
sfiimpprc12 <- FeatureImp$new(sfipred12, loss = auprcloss, compare = "difference", n.repetitions = 1000)

sfiimpprcplot12 <- plot(sfiimpprc12)

pdf(file.path(outdir, "2012FeatureImpsPRC.pdf"), width = 5, height = 4)
sfiimpprcplot12 +
  ggtitle("Figure 1C") +
  scale_x_continuous(name = expression("Feature Importance (Loss: 1 - AU"[PRC]*")"), 
                     limits = c(-0.01, 0.22))+
  scale_y_discrete(name = NULL, labels = df12labs) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))

sfiimpprc12res <- sfiimpprc12$results
sfiimpprc12res
write.csv(sfiimpprc12res, file.path(outdir, "HRS 2012 SFI10 FeatureImpPRC.csv"), row.names = TRUE)

#######################
#SFI-8#
#######################
socimpprc12 <- FeatureImp$new(socpred12, loss = auprcloss, compare = "difference", n.repetitions = 1000)

socimpprcplot12 <- plot(socimpprc12)

socimpprcplot12 +
  ggtitle("Supplemental Figure 1C") +
  scale_x_continuous(name = expression("Feature Importance (Loss: 1 - AU"[PRC]*")"), 
                     limits = c(-0.01, 0.045), 
                     breaks = c(0.00, 0.01, 0.02, 0.03, 0.04)) +
  scale_y_discrete(name = NULL, labels = df12labs,) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 14), axis.title = element_text(size = 13),      
    axis.text = element_text(size = 13))

dev.off()

socimpprc12res <- socimpprc12$results
socimpprc12res
write.csv(socimpprc12res, file.path(outdir, "HRS 2012 SFI8 FeatureImpPRC.csv"), row.names = TRUE)
```


# Item correlations with age: 2012 cohort#

```{r}
#######################
#correlations of social items with age as coded by HRS#
#######################
#need to recode variables first so that higher value#
#direction associated with mortality in SFI#
#######################
#age#
#no change#

#neighborhood clean?#
#no change#
table(df12oc$neighborhood)

#financial control#
#REVERSE 0-10#
table(df12oc$financial_control)
df12oc$financial_control <- 10 - df12oc$financial_control
table(df12oc$financial_control)

#has children/freq. seeing#
#no change#
table(df12oc$freq_children)

#activities w/ youth#
#no change#
table(df12oc$activities_youth)

#work#
#code (5) to (0), then REVERSE#
df12oc$work[df12oc$work == 5] <- 0
table(df12oc$work)
df12oc$work <- 1 - df12oc$work
table(df12oc$work)

#volunteer#
#no change#
table(df12oc$volunteering)

#feel isolated#
#REVERSE 0-2#
table(df12oc$feel_isolated)
df12oc$feel_isolated <- 2 - df12oc$feel_isolated
table(df12oc$feel_isolated)

#treated courtesy/respect#
#REVERSE 1-6#
table(df12oc$courtresp)
df12oc$courtresp <- 6 - df12oc$courtresp
table(df12oc$courtresp)

#######################
#correlations of social#
#items with age raw HRS#
#######################
ogcordfspear12 <- df12oc[c("age", "neighborhood", "financial_control", 
                    "freq_children", "activities_youth", "volunteering", 
                    "feel_isolated", "courtresp")]
view(ogcordfspear12)
ogcordfpear12 <- df12oc[c("age", "work")]
view(ogcordfpear12)

ogcordatpear12 <- cor(ogcordfpear12, method = "pearson")
ogcordatspear12 <- cor(ogcordfspear12, method = "spearman")
subsetogpear12 <- ogcordatpear12["age", ,drop = FALSE]
subsetogspear12 <- ogcordatspear12["age", ,drop = FALSE]

ogcord12 <- cbind(subsetogpear12, subsetogspear12)
ogcordat12 <- as.data.frame(ogcord12)
ogcordat12 <- ogcordat12[, c("age", "age", "freq_children", "activities_youth", "volunteering", 
                          "feel_isolated", "neighborhood", "financial_control", "courtresp", 
                          "work")]
ogcordat12 <- ogcordat12[, c(-2)]
ogcordat12 <- as.matrix(ogcordat12)
colnames(ogcordat12) <- df12labs[colnames(ogcordat12)]
rownames(ogcordat12) <- df12labs[rownames(ogcordat12)]


#visualize#
png(file.path(outdir, "2012agesocialitemscors.png"), width = 9, height = 9, units = "in", res = 600)
socogplot <- corrplot(t(ogcordat12), method = "color", addCoef.col = "#000000", 
                     col = colorRampPalette(c("#6A4C93", "white", "#27933E"))(200),  
                     col.lim = c(-1, 1), tl.cex = 1.4, tl.col = "grey30", 
                     cl.pos = 'n' , number.cex = 1)
dev.off()

#######################
#following up on positive correlations#
#######################
#activities w/ youth#
cor.test(df12oc$age, df12oc$activities_youth, alternative = "greater", 
         method = "spearman", exact = FALSE, conf.level = 0.95)

#volunteering#
cor.test(df12oc$age, df12oc$volunteering, alternative = "greater", 
         method = "spearman", exact = FALSE, conf.level = 0.95)

#work#
cor.test(df12oc$age, df12oc$work, alternative = "greater", 
         method = "pearson", conf.level = 0.95)

#feel isolated#
cor.test(df12oc$age, df12oc$feel_isolated, alternative = "greater", 
         method = "spearman", exact = FALSE, conf.level = 0.95)
```